---
title: "A First R Tutorial"
author: "Simon Hans Edasi"
date: "4/13/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Q SCI 381 Introduction to Probability and Statistics: A First R tutorial

\begin{centering}
WELCOME TO THE FIRST R TUTORIAL!
\end{centering}
 In these tutorials we are going to learn to 
 use a powerful, and currently, the most commonly used
 software package, R. This package will significantly
 decrease the tedious calculations that once were a
 common part of statistical analysis.  After learning
 how to use R, you will be able to use it in your other 
 courses and mention the ability to use it on your resume!
 
  NOTE: We are not going to work directly in R. We 
 will be using an interface, R Markdown. An R Markdown
 file has an *.Rmd extension wile an R file has an
 *.R extension in its title.  R Markdown allows us
 to type and run R script while also making printing 
 and sharing all of the output of your work with others
 easier than in R.
 Using R Markdown permits printing all of the output of
 the scripts INCLUDING any plots that are created. In R,
 this requires much more work.  Later in this tutorial, 
 we will learn how to include plots in R.  
 You will see why R Markdown is the preferred way of working
 with what you do in R if you wish to print out the results.
 
 In these tutorials and the assignments, you will read 
 a reference to the symbol "#" representing a comment
 or nonexecutable statement in R. Our R Markdown 
 files have been written so that the "#" is not 
 necessary for adding comment statements. What you 
 are now reading is a comment statement in R Markdown.
 You do not need to precede these lines with a "#".
 You may add lines like these as you like.
 
 In R, comments are liberally used to explain
 your code and make it readable.
 You would be amazed by what you
 can forget in a short period of
 time after you have written
 some code. Don't assume that 
 you will remember anything about
 the code when you look at it later.

 
 In these tutorials the following will give you an area 
 to place your code (R script). Type your script between
 the ```{r} and the closing ```as shown below.
 If you need more space for writing code,that is easily
 done by using the "Enter" on your keyboard.
 


```{r}

# Enter your code (R script) in this space.  Note that I included a "#".
# See below for the meaning of this symbol.


```

 If you want to add a comment statement in an area where you
 write code in R Markdown, then be sure to precede the statement
 with a "#" symbol. This is shown in the lines below.


```{r}
# This is a comment statement. I also used it in the code "chunk" above.
# Notice the three icons to the upper right of this block.  Hover your cursor over
#each one to see what they allow you to do.


```

 An important comment before we begin these
 tutorials: Save your work often!
  
 It is at the very least frustrating to work
 for a long time to figure out how to write 
 some instructions only to mistakenly
 hit the wrong key or have the software package
 "bomb" or stop working (and that can and 
 does happen!) and then have to redo everything. Let's begin.

 Please keep something in mind as you embark 
 learning R (or any software), especially if you
 have never written a computer command before in
 as software package: be patient. Take it slow, but keep at it!
 
 We will be here to help you when you need it, but
 please try to work things out on your own first.  
 If you do that you will gain confidence and skill.
 You can always refer back to these tutorials
 and BC&P throughout the course.

 As our guide to learning the basics of R, 
 we are going to use: 

    A.  P. Beckerman, D. Z. Childs, and O. L. Petchey. 
     2017 (2nd Edition). Getting Started with R:
     An Introduction for Biologists. Oxford.

 We will refer to this text as BC&P.

 I hope that you will find it supportive for a
 new learner and "user friendly".  The authors are
 British, so the word "maths" is not a misprint of 
 "math".  The same is true of "colour".
 
 We will not use all of the text.

 Many people have successfully learned to use
 R before you and you can do it also!

 Our TAs and I are here to support you.  We 
 have your back.  Try things on your own first, but
 when you just don't get it, please ask us for help.
 We will not use all of the text.

 Before doing this tutorial, please
 read pages iv - 8 (up to section 1.5) 
 and pages 219 - 221 in BC&P and you will be
 asked to reread them after you have completed
 A Third R Tutorial.

 Pages 219 - 221 are closing remarks, but it
 may help you to read them now before you start
 this tutorial.  Reading these pages and 
 completing these three tutorials will help you 
 get started. The third tutorial will focus on
 creating useful plots (graphs) of data.  
 This will be central to our study of statistics.
 
 Feel free to read everything you like in BC&P
 beyond what is covered in these tutorials,
 especially if you think or know that you will
 be using R in your work or other courses. 
 R is used in Q SCI 482 and 483 as well as many 
 courses in BIOL, SAFS, SEFS, and SMEA.
 In this course, we will only use a samll amount
 of what R can really do.

 As you read BC&P do it actively, i.e., with RStudio
 open on your computer.  Use that computer to enter
 and RUN the scripts (code) in this
 tutorial and in BC&P.  That active
 engagement will not only make learning the basics of
 R easier BUT more interesting!  

 This tutorial will introduce a few  other functions  
 that are resident in R for your use in doing
 operations more easily, but not mentioned in BC&P 
 so that you will have a few more tools
 available in you R tool box.
 
 As you work along with the text in BC&P, learn
 the good habits of writing and documenting your
 script. The text makes a great effort to point 
 these out. Learning to do this now will permanently
 pay big dividends to you.

 Welcome to RStudio. The flexible
 user interface for R.  It consists of four 
 panes with multiple tabs containing various 
 options. We will learn to use these panes
 and many of the options.

 R is the most widely used statisitcal software
 today.  It is open source and free to download.
 But, here is a lesson in basic economics.
 It is free to download, but it is not really free.

 You have to take your time to download it, set it up,
 learn its syntax and commands, and how to use it productively.
 That all takes time, and time is the most important 
 and expensive asset we have.  It can never be replaced or
 renewed. Economists call these costs opportunity costs.
 By this they mean that the cost of anything is your 
 next most favored (preferred) alternative that you gave
 up to get what you selected.  In this case,
 that is the time you could have spent doing
 something else.  But, learning some R will
 help you later, so it will be worth your time.
 This ends our economics lesson. :-)

 Many expert users have created functions that 
 allow us to do data manipulation, statisitcal
 analyses, and plot our data and results in 
 publication quality.  If you wish, you can also
 learn to write your own fucntions to carry out 
 your own analyses.  Our use of R will be limited
 in Q SCI 381.  We will not fully explore 
 its capabilities, but now that you have it
 you can if you wish.  We will be happy to help you.
 Since R is so widely used, the more you learn
 to do with it, the more career opportunities
 may be opened to you.

 Before we start, go to the toolbar above,
 click on the "Help" tab (this may be in the lower right depending on 
 your version of RStudio), then click R Help.
 Look at the Help tab that opens in the lower
 right pane of RStudio.  Look at all of the
 built in help that is available to you;
 including a "Learning R Online." This help
 always available to you.  You really are not alone
 if you are working on these tutorials or the
 course assignments even at at 3:00 a.m.  Please 
 have the patience to go to the help files and
 use them.  you may also use natural language
 searches in Google for a wealth of worked examples
 to do almost anything you wish to do.  Really!

 Look at the R Cheat Sheets.  Several of these have
 already been downloaded to our course Canvas website.

 Finally, click on the "Files" tab In the lower right 
 pane of RStudio, then click the little house with the
 red roof.  What do you get?
 Yes, help is just a click away.

 The developers of R have tried to provide as much
 information as possible to make learning and using
 R as easy as possible.

 Liberally use white space 
 between code segments for ease 
 of reading and understanding. 

 Don't forget to
 save your scripts (R Code) 
 frequently.  As  BC&P point out,
 the two most important things
 in R is your data and your script, 
 which is the code you write to 
 have R do something with the data.  
 Keep both them safe! Save often!!

 Rstudio has great flexibility 
 programmed into it. It will often 
 prompt you if you have made an error
 in typing a function or command.
 It has all of the usability of
 word processing software. It has its
 own text editor including "undo", "redo,
 "cut", "paste", and "copy" operations in
 the Edit tab above in this pane.
 Warning!  RStudio does not have a very 
 sophisticated spell check and no 
 auto correct feature except for its 
 built-in commands.  That is tough when
 you have become use to using those two 
 features, as I have.
 
 So, check your spelling in your comments!
 I know that from personal experience :-).

 You may also search for words or phrases
 in your code by clicking the magnifying
 glass in the tool bar above.

 Look around each of the panels and 
 check out what options the tabs
 provide.  If you lose the 4 panel
 display, just click 
 "View > Panels > Show All Panes".

 Always start by clearing the memory
 of your computer's microprocessor.
 
 R accomplishes this clearing of memory
 with the command (object or function) "rm()" 
 in R "rm()" stands for remove 
 what is in the "()".  This
 command will help remove anything extraneous
 and unwanted by you in the memory 
 of your computer's microprocessor, e.g.,
 the use of a variable name declared (defined)
 in some other work that you may have been
 doing previously that you now want to use for 
 something else.

 In the space designated for code Below this comment, enter 
 the following (type it or cut and paste it):

     rm(list=ls())  

 Please execute (run) it.
 Does anything happen in the Console?


```{r}
rm(list=ls())
```

 Statements like "rm()" are built-in coded 
 functions (or commands if you like)in R.
 Many people have contributed functions for use
 in R and they are available for our use. 
 The statements in the "()" of a function are its
 arguments. They control what the function does.
 We will learn to use several of these for the 
 built-in functions in R. R provides a way to
 find out how to use its functions and what
 each function's arguments do.  The names
 of objects (and in R a function is an object)
 are case sensitive (capital and lower case matter).
 More about this later in this tutorial.

 One comment "()" are not the only delimiters used
 in R's syntax.  You have to be careful.  Both "{}"
 and "[]" are used.  Things like that are common in
 all software, so it will be good to encounter it
 here and it will not come as surprise if you
 use other software programs.

 Learning a little bit of R will be valuable
 in other courses. That will prove
 to be a benefit to you by extending what 
 are able to.  This knowledge with expand your
 career opportunities and flexibility.

 Set the working directory (where you will save you data and code)
 below if you like. Use the following code:
 
     setwd("C:/Your/Path/Here")


```{r}
setwd("/home/simonhans/coding/qsci/hw1")
```

 An example of getting help
 with functions, objects, and commands.
 In a line below these comments, type

     ?log  

 What appears in the lower
 right pane of RStudio after you 
 click the "Help" tab?  
 
 **A detailed description of the logarithm mathematics and examples of how they are used in R.**
 
 What base does this 
 log function use? 
 
 **This function by default uses base exp(1), but can be defined with any number base. log10() is used for base 10 and log2() is used for binary bases.**
 
 Look at what arguments
 this functions can use.   This is the place
 we can look for help within RStudio.  
 Of course, you may also
 use Google and printed books 
 to get  help.  The point is that 
 it may take some work, but there
 are many places to get help.  You are
 not strictly on your own!


```{r}
?log
```

 Rstudio can help you complete
 statements. In an open line below
 type 

    sub 

 
```{r}
sub
```

 Press the Tab key
 on your keyboard.  What appears in
 the Console below?
 
**A list of potential fuctions that start with the letters 'sub'.**


 Remember when we started by
 executing the function 
 "rm(list=ls())"? 
 in a line below this comment
 type 

     ?ls()
 and read what appears 
 under the Help tab in the lower
 right pane.  By executing 
 "rm(list=ls())", what is it that 
 you have asked R to remove?

**That code removes all visible objects and variables from the current environment. The ls() function returns a list of vectors, and by including it as an argument we tell Rstudio to remove everything on that list.**

```{r}
?ls()
```



 After typing a line of code, you can
 have RStudio run (execute it) by typing
 Ctrl-Enter together (on a PC).  
 Let's try this in an open line below this comment,

 type 

    ?sin  

```{r}
?sin
```

 When you do this, you will notice that RStudio
 presents you with several possible choices.
 Click on the "sin" choice.  See what appears 
 in help in the lower right pane.  
 We get some help. Great!
 Pretty cool, huh?


 Note that highlighting the line or lines of 
 code and clicking Run from the menu above 
 will also execute the code.  At this point,
 please click the Code tab above and review
 the options available to you.

 I will warn you that the people who
 write the code for these operations,
 objects, or functions may use different
 ways of documenting them.
 There may not be consistency in the documentation.

 Another problem is that those who write the code for
 these functions get, within certain limitations 
 regarding special reserved words, to pick the
 words they use.  Later that will cause us a bit
 of a problem when we want to find the "mode" of
 a data set.  Someone used that word for something else
 and it can not be used by us to find the mode (most 
 frequent value in a data set).  Nothing really is free!

 We will learn how to create code that will
 permit us to have R perform
 many operations on data or other functions.

 This ability of R and Rstudio to create several
 lines of script (code) allows us to write a program that
 automates doing an entire statistical analysis of 
 our data.  To run all of the code, we simply 
 highlight all of it and click "Run" at the top
 of the script editor pane (upper left pane).  
 At this point, explore what options appear in the
 Code tab above.  Try some of them.
 Remember how to restore RStudio to its 4 pane
 view if you should change the view.

 Now type 

     sin  

```{r}
# sin()
```

 click the Tab key on
 your keyboard.  What was added after
 the sin?
 
 **()**
 Reenter "sin" in the space below and run the code.

 In the parentheses that appear, type

     pi/2 

 (all lower case), then click Ctrl-Enter or the green arrow in the upper right corner.
 What appears in the Console?
 
 **1**
 
```{r}
sin(pi/2)
```

 In the upper right pane, open the 
 History pane.  What shows up there when you
 executed sin(pi/2)?
 
**sin(pi/2)**

 CONGRATULATIONS!! You have
 just gotten R to calculate something for you!
 That was pretty easy.  From here on we will 
 just build on this and add more capabilities.

 Before we move on to working our
 way through guided instructions in BC&P,
 let's get R to do some computations for us.

 R can work like a giant calculator.
 A calculator limit only by the
 capabilities of your computer's
 microprocessor.

 In the two lines below this comment, 
 you will notice that the use of parentheses 
 and the order of algebraic operations is 
 important. It is always important. Please  
 do not ever forget this valuable lesson.
 The results of the two lines below
 are different!  Please do not forget your
 algebra!!

 Type these two lines:

    1+2*(3+4)
    1+2*(3)+4

 in two lines below this comment
 and see what results are returned by R
 in the Console.

```{r}
1+2*(3+4)

1+2*(3)+4
```


 What happens if we forget the "*"?
 Try it for yourself. Retype the 
 following on the lines below this comment: 

    1+2(3+4)
    1+2(3)+4

If you are looking at the key, you will need to delete the * below.    
```{r}
# 1+2(3+4)
# 
# 1+2(3)+4
```

 What appears in the Console?
 
 **Error: attempt to apply non-function**
 
 Are the answers the same?
 
 **no**
 
 Do the indicated algebra by hand
 and make sure you understand
 the answers appeared as they did.
 This is important for you to do
 for yourself.  Please do it.

 This is pretty cool because we already
 know how to do this from algebra.  This 
 is just the order of operations (remember the 
 mnemonic "Please Excuse My Dear Aunt Sally").

 Now try each of the following 
 calculations:

    log(4^3+3^(2+1))
    sqrt((4+3)+(2+1))
    ((1+2)/(3+4))^2

 Keep track of what appears in the Console.

```{r}
log(4^3+3^(2+1))
```

```{r}
sqrt((4+3)+(2+1))
```

```{r}
((1+2)/(3+4))^2
```
 Since R is case sensitive (lower and
 upper cases are important), in a line below this
 comment type: 

   SQRT((4+3)+(2+1))  

 What appears in the Console?
 
 **Error : could not find function "SQRT"**

```{r}
# SQRT((4+3)+(2+1))
```

 On the lines below, please create a complex 
 expression of your own to evaluate.  Add
 comments to help you remember what the code does.

```{r}
# note this can be any expression
# Logarithm of sum of two trig functions of exponentials whose arguments are gibberish numbers

log(sin(exp((pi/4 * 72)/14) - 89999) + cos(exp((pi/4 * 72)/14) - 89999))

```

 Let's look at an important concept that BC&P 
 will also cover.  The assignment of values.
 Above we calculated the "log()".  Once calculated,
 the value returned is no longer available for us
 to use in the rest of our code.  What do we do to 
 keep these values for later use?  

 We assign them to a variable. 
 We tell R that we are assigning
 something (including another object) to a variable
 by using either "<-" or "=" as follows:
 x<-log() or x=log().  
 Statements like this are read
 "x gets (or is assigned) to have the value log()", or
 whatever appears on the 
 right hand side (RHS) of the <- or =.  
 
 Notice that "=" is no,longer the equal sign 
 from algebra.  Instead equal sign is 
 reserved for something special. 
 The equal sign you are used to from algebra
 becomes "==".  This formalism is common
 in most software and you should become familiar 
 with.  It is not hard to get used to. We will have
 an opportunity to use "=="  as we use R later in
 the course.

 Let's assigning a value to a variable. We will
 try this in three ways using the "log()" function.

 Enter and execute the following:

    answer1<-log(exp(1))
    answer1
    answer2=log(exp(1))
    answer2
    answer3=log(exp(1), base=10)
    answer3

 What appears in the Console?
 
 **> answer1<-log(exp(1)) \ > answer1  \ [1] 1 \ > answer2=log(exp(1)) \ > answer2 \ [1] 1\ > answer3=log(exp(1), base=10) \ > answer3 \  [1] 0.4342945**

```{r}
answer1<-log(exp(1))
answer1
answer2=log(exp(1))
answer2
answer3=log(exp(1), base=10)
answer3
```

 Another way to view the objects
 answer1, answer2, and answer3 is by 
 typing "print()".  Let's try this, but
 don't forget to highlight these lines of 
 code and either click "Run" from the
 above menu or hit the Crtl-Enter keys
 simultaneously.  What did you get in
 the Console?
 
 **> print(answer1)[1] 1> print(answer2)[1] 1> print(answer3)[1] 0.4342945**
 

```{r}
print(answer1)
print(answer2)
print(answer3)
```

 When you do one of these actions,
 your R script is sent to the Console
 where R carries out the commands in your
 script.

 Now we let's look at one of the great
 advantages of assigning things to objects.
 We are going to manipulate or use the "answer" 
 objects in other operations, e.g., on a
 line below type and execute the command:
    answer1*10 
 What did you get as an answer in the Console?

**> answer1*10  [1] 10**
```{r}
answer1*10
```

 Did you notice?  We could simply have R do some
 manipulation on answer1.  Please try your own
 manipulation of answer1 in the lines below this comment.
 It is just like algebra! Cool, huh?

 Let's do one more famous operation with 
 objects. It is famous because every
 person learning programming has done it.

 Try typing this: 

     MyName<-"Enter your name here"

 on a line below these comments.  Make sure to
 put your name between the " ".
 Recall that this says that the object (variable)
 MyName gets (is assigned to)the value Your name.
 Follow this code by "print(MyName)" and "MyName".
 Did you get the same result in the Console?
 
 **yes**
 

```{r}
MyName <-"Simon Hans Edasi"
print(MyName)
```

 Let's look at how to list and remove objects
 in our workspace. First, highlight all of
 code (script) above and click Run from the 
 menu above.  Let to code run.  After the code 
 finishes running, type in a line below this comment 
 and then execute the function (command) "ls()".
 You know what "ls" stands for.
 What appears in the Console below?  How were
 these objects created?  We created them with
 our assignments further above in this tutorial.

 **What appears in the console is the list of objects we have created with previous R code**


```{r}
ls()

```

 Now let's remove answer2 from our workspace.
 On lines below this comment type and execute
 the following:
    rm(answer2)
    ls()
 What appears in the Console?


```{r}
rm(answer2)
ls()
```

 Now we will remove all of the objects
 from our workspace.  Can you think of
 how to do this?  You have done it
 before in this tutorial.
 On lines below this comment type
   rm(list=ls()) .
 What happened?  Where do you look
 to answer that question?
 
 **look in the environments tab for our variables and see that it is empty**


```{r}
rm(list = ls())
```

 So how do we get data into R?
 That would seem to be an important question
 since in Q SCI 381 we wish to learn 
 how to explore and analyze data.  More about this
 will be discussed in Chapter 2 of BC&P.  For now
 we will consider some simple ways of doing this.

 We can enter data by creating a list or vector 
 of data values.  We do this by assigning the data
 values to a variable using the concatenation 
 function in R.  Concatenation means joining items
 end-to-end. This effective creates a vector.
 The syntax is "x<-c()".  the data are 
 entered within the "()".

 In a line below this comment, type and execute
 "?c()".  Remember how to execute this line of
 code.  What shows up in the Help tab in the
 lower right pane? 
 
 **Documentation for the c() function, which combines values into a vector or list**
 
 Help is useful and always
 available.


```{r}

?c()

```

 Importantly, R has several named functions.  These
 names are reserved and We can not use them as names for
 our variables.  R will warn you with an error message
 if you do violate that constraint.

 The data below are the annual flow measurements of the
 river Nile at Asswan from 1871 - 1970, in 
 trillions of cubic meters 
 (i.e., trillions of cubic meters per year). The data
 come from Cobb, G. W. Biometrika. 65(2): 243 - 251.
 They are listed in chronological order.

 1120, 1160, 963, 1210, 1160, 1160, 813, 1230,
 1370, 1140, 995, 935, 1110, 994, 1020, 960,
 1180, 799, 958, 1140, 1100, 1210, 1150,
 1250, 1260, 1220, 1030, 1160, 774, 840, 874,
 694, 940, 833, 701, 916, 692, 1160, 1050, 
 969, 831, 726, 456, 824, 702, 1120, 1100,
 832, 764, 821, 768, 845, 864, 862, 698, 845,
 744, 796, 1040, 759, 781, 865, 845, 944, 984,
 897, 822, 1010, 771, 676, 649, 846, 812, 742,
 801, 1040, 860, 848, 890, 744, 749, 838, 
 1050, 918, 986, 797, 923, 975, 815, 1020, 
 906, 901, 1170, 912, 746, 919, 718, 714, 740

 In a line below this comment, enter this data
 using the "c()" function.  Call your vector
 RiverNileFlow.  Do this the easy way by
 cutting and pasting the data above. Copy
 the entire list and paste it 
 within the "()". You
 can cut and paste the variable's (vector's) 
 name as well!  Nice, huh?

 Notice what shows up in the Console in
 RStudio.

```{r}

RiverNileFlow<-c( 1120, 1160, 963, 1210, 1160, 1160, 813, 1230,
 1370, 1140, 995, 935, 1110, 994, 1020, 960,
 1180, 799, 958, 1140, 1100, 1210, 1150,
 1250, 1260, 1220, 1030, 1160, 774, 840, 874,
 694, 940, 833, 701, 916, 692, 1160, 1050, 
 969, 831, 726, 456, 824, 702, 1120, 1100,
 832, 764, 821, 768, 845, 864, 862, 698, 845,
 744, 796, 1040, 759, 781, 865, 845, 944, 984,
 897, 822, 1010, 771, 676, 649, 846, 812, 742,
 801, 1040, 860, 848, 890, 744, 749, 838, 
 1050, 918, 986, 797, 923, 975, 815, 1020, 
 906, 901, 1170, 912, 746, 919, 718, 714, 740)




```

 In a line below this comment type or cut and
 paste the variable's name, i.e., RiverNileFlow.
 Execute this variable (object).  
 What shows up in the Console of RStudio?
 
 **the list of numbers contained in the object named RiverNileFlow**
 
```{r}
RiverNileFlow



```

 Now that we have our data in R, only our creativity
 and what we know about R limits what we can do with it.
 We can manipulate it; explore it, by plotting it,
 calculating measures of its central tendency 
 (mean, median) and dispersion (standard deviation); and 
 analyze it.  

 In short, we can use the full capabilities 
 R's built-in functions.  We can even write our own
 programs to analyze our data.  R can do a great deal.
 Let me repeat, You are only limited by your creativity
 and what you know about using R.  

 Since we took the time to enter this data,
 Let's do something easy with it.  Why not?
 
 Recall that our variable is a vector (an ordered
 collection of data).  How many values do we have?
 What is the first value?  What is the last value?
 This data set is small, and we could answer these
 questions by just "eyeballing" (looking at) the data.
 But, would you want to do that if you had 10,000
 data items?  I wouldn't.

 Guess what, the developers of R and those that 
 continually contribute functions to its growing inventory
 of capabilities thought about that and R has functions that
 will answer those question for us.

 We are going to look at using the functions:

    RiverNileFlow
    length(RiverNileFlow)
    RiverNileFlow[1] 
    RiverNileFlow[length(RiverNileFlow)].

 Consider the function RiverNileFlow[1]. 
 Notice the "[]" synatx. This function requires that
 the "[]", and only the "[]", be used with this function.
 Learning syntax is important in all software, and 
 it is not consistent among software products.  Sorry!
 The "1" is an index indicating (requesting, referencing,
 or pointing to) the first element in
 the vector (variable) we have named RiverNileFlow.
 
 Knowing this, we can see that length(RiverNileFlow)
 must represent an index pointing to an element in
 the vector RiverNileFlow.  What value do
 you think it is referencing?  The function "length()"
 produces the number of elements in the object 
 (vector or other) within the "()".  So, 
 RiverNileFlow[length(RiverNileFlow)] returns the last value
 in our variable (or vector) RiverNileFlow.  That is 
 pretty cool!  R will do that for use in a data 
 set of millions of data points!  And it does it in seconds!  

 Before going on, look again at that last function.
 That is a lot of typing and we might
 mistype the variable name. That is frustrating.
 You may also notice that the name of the variable
 is very informative.  The lesson is, you have balance
 being informative and work of typing.  Personally, 
 I have always weighted the balance toward informative. 
 After all, I can use cut and paste!  

 Don't forget about the "undo" operation available
 in the Edit tab on the tool bar in case you make a mistake. 
 I have used that more than once!!

 In the lines below this comment, type the name 
 of our variable (use cut and paste). Execute that line of code
 and see what appears in the Console in RStudio.
 R has kept the data available to us.  Very thoughtful!
 
```{r}
RiverNileFlow

```

 After you do that cut and past the four functions 
 listed above, execute all four lines of code and look
 at what appears in the Console.  
 
```{r}
 
RiverNileFlow
length(RiverNileFlow)
RiverNileFlow[1] 
RiverNileFlow[length(RiverNileFlow)]



```

 The results (output) of the code we execute in R appears 
 either in that Console or, if it is a plot,
 in the Plots tab in the lower right pane. 
 The results in those panes can be cut and pasted 
 into documents (e.g., WORD documents).  RStudio will
 also create pdfs.  We will use the cut and paste 
 operation in the class assignments.


 Now run (execute) the following
 two lines of code below this comment:

    LastValue<-RiverNileFlow[length(RiverNileFlow)]
    LastValue

 After executing these lines look in the Console
 of RStudio and see what appears.

```{r}
LastValue<-RiverNileFlow[length(RiverNileFlow)]
LastValue


```


 If we have several vectors (several data sets)
 R allows us to perform algebraic, logical, and 
 statistical operations on them.

 R will also let us work with matrices, i.e., data sets
 consisting of rows and columns of numbers. 
 If you do not know what a matrix is, 
 just imagine a spreadsheet with cells.
 Each cell has a row and column coordinate that
 uniquely identifies it. 
 Think about data on 100,000 people
 (or 100,000 stocks on a stock exchange). In the first case, 
 there is one row in the matrix for each person, 
 and stored in each column is a value of a given 
 characteristic of interest for that person,
 e. g., information about the person's
 age, sex, height, weight, eye color, etc. Each column
 would have a heading describing the characteristic.
 In the example of a stock, these characteristics
 might be the price per share of the stock,
 the number of shares outstanding, the last five years of
 revenue for the company who issued the stock, 
 and so forth.

 We will be working with data arranged
 like that later in this course.
 We will want to enter the data into RStudio, 
 manipulate it, explore it, and analysze it using R.
 R will allow us to do all of that.

 Finally, here is one more way of entering data.
 Let's enter the integers 1 through 10 in two ways,
 one of which you already learned.  For the first way
 will use "c()" and the second way will use "seq()".
 First, we should find out about the seq() function.
 How do we do that?  Type and execute it on a line
 below this comment.

    ?seq()

 What does the documentation about "seq()" tell us
 about this function.  Among other things, it tells
 us what goes in the "()"; the arguments of the function.
 The arguments cause the function what we want from it
 or what we want it to do.
 Note that seq() stands for "sequence generation".

```{r}

x<-c(seq(1, 10, 1))
y<-c(seq(10,1,-1))


```

 We will be working with several functions
 that have parameters.  The "plot()"
 function will be a very useful function to us.
 The arguments of plot() make the function very flexible.
 Take some time and check out the documentation on "plot()".

```{r}
?plot()

```

 Enter the following lines of code below this comment:

    MyNums1<-c(1,2,3,4,5,6,7,8,9)
    MyNums1
    MyNums2=seq(from = 1, to = 10,by = 1)
    MyNums2
    MyNums3<-seq(1,10,1)
    MyNums3

 A note on syntax.  Carefully notice the space in
 MyNums2=seq(from = 1, to = 10,by = 1)
 before and after the "=".  The "=" is interpreted
 as LHS "gets or takes the value" what is on the RHS.

 What appears in the Console?
 Is there anything special about MyNums2 and MyNums3?

```{r}
MyNums1<-c(1,2,3,4,5,6,7,8,9)
MyNums1
MyNums2=seq(from = 1, to = 10,by = 1)
MyNums2
MyNums3<-seq(1,10,1)
MyNums3







```

 The result of MyNums2 and MyNums3 are the same.  That
 is because the two statements
 seq(from = 1, to = 10,by = 1) and
 seq(1,10,1) are equivalent.  This will often be the case
 in R, but it is best to check the documentation to be
 sure that is the case.

 Now on the lines below this comment,
 type the following code that manipulates
 the object MyNums3 and see what you
 get:

    MyNums3TimesTwo=(MyNums3)*2
    MyNums3Squared=MyNums3^2

 We want to see our results, so we need
 to result that results be displayed.  Do this
 as follows:

    MyNums3
    MyNums3TimesTwo
    MyNums3Squared

```{r}
MyNums3TimesTwo=(MyNums3)*2
MyNums3Squared=MyNums3^2


MyNums3
MyNums3TimesTwo
MyNums3Squared


```

 For another example of what R can do,
 enter the following lines of code
 (or cut and paste them) on the lines below:

    library (ggplot2)
    x<-seq(-10,10,0.1)
    y=x^2
    qplot(x,y,geom='line')

 What appears in the lower right pane?\
 
 Note that you may first need to run the following: 
 
 install.packages("ggplot2")
 
 in order to get the above code to run. Once you've installed a package, you will need to    activate it by using the library command.

```{r}
# install.packages("ggplot2")
library(ggplot2)
x<-seq(-10,10,0.1)
y=x^2
qplot(x,y,geom='line')




```

 You can export plots to incorporate into the  
 assignment you turn in.  You may have to experiment
 with how to do this with the software you are using.
 We will try to help you with this, but you can always 
 save the plot to your desktop and then print it as 
 a separate page and attach it to your solutions.

 If you are using MS WORD, there are at least two ways 
 can do this: 
 
 [1] You can export this plot for pasting into a WORD document
     as a PDF.  In the lower right pane of RStudio click "Export" 
     then select "Save Plot as PDF".  Select the appropriate PDF
     size for the plot. Select a good size (4.00 x 6.00 may be a good
     place to start).  Check your plot in "Preview".  When you 
     think that you have a good plot, select a file name and a 
     directory in which to save it.  In WORD, select the Insert tab.
     In the Text group, click Object.  In the Object dialog box,
     click Create fron File.  Find the file using Browse, then 
     click Ok.
 
 [2] You can also save the plot as a metafile. To do this, 
     select "Export", then "copy to Clipboard", 
     then select "Copy as Metafile".
     Make the plot have a height of about 600,
     but you should experiment with this setting to 
     get a good quality graph that clearly displays
     the relationships in the plot.
     
 You have learned to use R to do a good deal of work.
 R will be our "work horse". 
 We use just enough of R's capabilities
 to do the tedious work we need do to implement
 the procedures we will learn.

 Since the class is only an introduction
 to probability and statistics, We will not learn 
 R in depth.

 The class assignments will guide you through
 what R you will need. But as you go through the 
 please take the time to notice just how much R
 can do, and how much you will have learned to 
 use R to do.  You may even come to think that R
 is pretty useful and worth learning more about.

 The BC&P text is "user friendly" and will be 
 a good beginning reference for the class.
 Please actively read it.  By actively reading,
 I mean reading the text with RStudio open on your
 computer and doing the examples as you read the text.
 That active reading is what we will do in the 
 next R tutorial.

 Learning R and the power it can give you to
 do analysis in depth will be left to other 
 courses given on the UW campus or online. 
 However, while you are in 381 please feel free to
 learn as much as you wish about using R.
 Our TAs and I will be happy to help you.
 
 Now, take the time to reflect on 
 just how far you have progressed in
 starting to use R.  No matter whatever tedious
 work We use R to do for us, that is all R can do for us. 
 It can not ever substitute for our thinking! It
 will be your ability to think that will make you
 valuable in your careers.
 
 It always will be up to us to figure out
 what data we need, how to gather it, 
 what exploration of it we should do, 
 what assumptions are appropriate,
 what analyses are appropriate, and finally
 what does the analysis mean in the 
 context of our problem. 

 R, like all other software, was programmed 
 by people to perform certain specified tasks.
 It does not know if what you are asking it
 to do is legitimate, if it meets the underlying 
 assumptions that validate what you are using
 it to do.  It will gladly violate any and all 
 assumptions as this violation does not contradict
 what it is programmed to do.  It will treat
 integers as real numbers.  It will work with
 invalid data.  That, is where you come.  
 You must know that what you are doing is correct.
 You must know what you expect to come out of
 the software, and that what you see makes sense.

 In class we will spend time learning the
 tedious tasks of the calculations that R is
 doing for us. This is absolutely necessary. 
 That will help you become a knowledgable
 user and interpreter of data
 and the results of its analysis because you
 will know what is actually being done with
 the data. But, once we have learned how to
 do the work, we will make R do most of the tedious  
 work of the analysis for us; saving most of our
 for time interpreting the results of that 
 analysis and drawing conclusions from it.
 
 Never forget that the results of 
 our efforts are ultimately going to be
 used by human beings to make decisions.
 we must do good work so those decisions 
 may be meaningful and well made.

 These ideas will be emphasized 
 throughout the rest of the course.

 Please reread pages iv - 8 (up to section 1.5).
 Then please read section 1.5 thru page 34.
 some of this new reading will now be review.  
 That repetition will be beneficial.
 It also serves to let you know that you 
 have already learned a good deal about using R.
 
 You will notice that 
 when we downloaded and opened RStudio, 
 all four panes appeared, so
 when you are asked to use the Console, just use it.
 There is no need to close the Script pane only
 to reopen it.
 
 Make sure to accompany your reading with
 using R to reproduce what you are reading.  That
 is active reading. (Please note that
 Figure 1.8,on page 20, was done using a Mac, not a PC.)

 Use the area below for your scripts as you
 work you way through section 1.5 thru page 34.
 Add more lines as you need them.

 Save your work to your desktop or some other file.
 DO NOT FORGET TO DO THIS!  You have done a lot of
 work. Do not lose it.





            CONGRATULATIONS!!! 

 Now that you have completed this tutorial.
 Please go on to "Second R Tutorial"



