---
title: "QSCI 482 Story 7"
author: "Simon Hans Edasi"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,out.width='250px', dpi=200)
library(car)

```

### 1a. Read the data into an R object and use the table() function to check that you have a balanced two-factor ANOVA design.

Balanced design will have same $n$ for each factor.
```{r}
df <- read.csv('7 Guineapigdata.csv')
df$supp <- as.factor(df$supp)
df$dose <- as.factor(df$dose)
table(df$supp)
table(df$dose)
```

### 1b. For the rest of the assignment (for grading simplicity), choose the dose to be Factor A and the supplement type to be Factor B. Test the underlying assumptions of the two-factor ANOVA, using Levene’s Test (short way around) for homogeneity of variances. Write down your conclusions.

Underlying assumptions:
 
1. Variances are the same

$H_0:$ Variances across factors are equal

$H_A:$ Variances across factors are not equal
```{r}
leveneTest(len ~ dose*supp,data = df)
```

We fail to reject the null hypothesis that variances are equal.

2. Data in each cell comes from a normally distributed population

```{r}
aov.res <- aov(len ~ dose*supp, data = df)
plot(aov.res)
```

Residuals appear normally distributed, we appear to be good here.

### 1c. Run a two-factor ANOVA on the data the short way around. Copy and paste the output table in R that includes the degrees of freedom, mean square, p-values etc. Interpret the resulting p-values in your own words.

```{r}
summary(aov.res)
```

### 1d. Provide the plot for standardized residuals, the Q-Q plot, and run a Shapiro-Wilks test (short-way around), to check whether the data are really normally distributed. Paste the plots into this document, and explain your conclusions about whether the data are normally distributed.

```{r}



```