---
title: "ANOVA Tutorial"
author: "Simon Hans Edasi"
output: pdf_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

   

                                       
  
## Section 5.6
 
## The Analysis of Variance (ANOVA)
 
 
 \vspace{1cm}                                      


  
 
  In the classic one-way ANOVA, the 
  explanatory variable is a categorical variable
  which has two or more levels, e.g., recall
  the Grazing variable in the compensation data
  set.  The Grazing variable had two levels, Grazed
  and Ungrazed.
  
\vspace{0.25cm}    
 
  We will follow along with BC&P.
  The dataset we are going to use centers on water fleas,
  Daphnia spp., and their parasites.
  
\vspace{0.25cm} 
 
  The question we are asking focuses on water flea growth
  rates and has two parts. 
  
\vspace{0.25cm} 
 
### 1. Generally, do parasites alter the growth rates of Daphnia?
      
\vspace{0.5cm}       
  
### 2. The data come from  a well-replicated and designed experiment, we can also ask whether ach of three parasites reduces growth, compared with a control, no parasite treatment.
 
  
\vspace{0.5cm} 

```{r}  
# Clear the memory
rm(list = ls())

  
# Load libraries
library(ggplot2)
library(dplyr)
library(readr)

```

  \vspace{0.5cm} 
  
  
### 3. Get the Daphniagrowth.csv dataset.The data is structured as a data frame.In this dataset, growth.rate is the dependent ( or response) variable and parasite is the independent (predictor) variable.
 
  To make the plotting of the results from the
  analysis of this data set easier, it has been
  modified from what BC&P created.  An 
  additional column has been added that provides
  an abbreviation for the genus and
  species of each parasite used in the study.
  Running View() will display the new structure
  of the data set. The new column is named
  "parasite".  the full taxonomic names appear
  in the column named "parasite.name".
 
   \vspace{0.5cm} 
   
```{r}
daphnia = read.csv('Daphniagrowth.csv')
```

```{r}
View(daphnia)
```
  \vspace{0.5cm} 
  
  Look at it using two functions: View(), and head()
  
\vspace{0.5cm} 
  
```{r echo = FALSE}
head(daphnia)

```

  \vspace{0.5cm} 
  
 
  Since this course is focused on use of statistics in the 
  biological sciences, it is reasonable to ask what kind of
  parasites are these?  You should find out what kind of
  organism Daphnia is.
 
   \vspace{0.25cm} 
   
  An internet search provides the following information:
  
\vspace{0.25cm} 
 
 1. Pasteuria ramosa is a gram-positive, 
       endospore-forming bacterium in the 
       Bacillus/Clostridia clade within Firmicutes. 
       It is an obligate pathogen of cladoceran crustaceans
       from the genus Daphnia. An established and widely
       used coevolutionary model of host-pathogen 
       interactions exists with P. ramosa and D. magna.
       
  \vspace{0.5cm}        
  
2. Pansporella perplexa is an amoeboid parasite. Look it
       up in relation to D. magna.
       
  \vspace{0.5cm}        
 
3. etschnikowia are single-celled fungal parasites
       of freshwater animals. It typically parasitises 
       crustaceans, including Daphnia, a genus of zooplankton. 
       M. bicuspidata has also been found to infect prawns
       and salmon.
       
  \vspace{0.5cm}        
       
       
  So, we have three different types of parasites; a bacterium,
  an amoeba, and a fungus.
 
  We can see that the data frame has four 
  variables, two of which we want to use for 
  figure making, growth.rate and parasite. 
  The other two, rep and parasite.name, indicate
  the replication in each treatment level and
  the full taxonomic name of each parasite,
  respectively.
 
  In Chapter 4, where we introduced the
  box-and-whisker plot as a quick and effective
  tool for viewing variation in a response
  variable as a function of a grouping, categorical
  variable? That's probably the start we want to make.
 
   \vspace{0.55cm}  

  
### 4. Create a box plot of the daphnia data.
  
\vspace{0.5cm} 
  
```{r echo = FALSE}
  
ggplot(daphnia, aes(x = parasite.name, y = growth.rate)) +
geom_boxplot() +
theme_bw()

```
  \vspace{0.5cm}   
 
### 5. Use one method to fix the graph for better understanding.
 
   \vspace{0.5cm} 
   
```{r}

ggplot(daphnia, aes(x = parasite.name, y = growth.rate)) +
geom_boxplot() +
theme_bw() +
coord_flip()

```

  \vspace{0.5cm}   
 
  Let's search for outliers and then plot
  the box plots. The script relies on an
  often used mathematical definition of
  what might be an outlier.  Ultimately,
  the determination of what constitutes an
  outlier rests with the investigator and their
  knowledge of their subject. A mathematical
  method unthinkingly employed is insufficient
  to determine an outlier.  The method is
  used here as a scanning tool to find
  potential outliers.
  
\vspace{0.5cm} 
  
  This script introduces
  a user created R function named "is_outlier".
  Such functions are highly useful for creating
  code that can be used many times within a
  larger R script. Doing this saves rewriting
  a similar piece of code many times within the 
  larger script.
  Learning how to create R functions is beyond
  what we will learn of R in this course,
  but perhaps you may now be able to 
  dissect the script to see how it works.
 
  \vspace{0.5cm} 
  
```{r}
is_outlier = function(x) {
  return(x < quantile(x,0.25) - 1.5 * IQR(x) |
           x > quantile(x, 0.75) + 1.5 * IQR(x))
}

daphnia %>%
  group_by(parasite) %>%
  mutate(outlier = ifelse(is_outlier(growth.rate), growth.rate, as.numeric(NA))) %>%
  ggplot(., aes(x = factor(parasite), y = growth.rate)) +
  geom_boxplot() +
  geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3)+
  geom_point(size = 1, colour = 'black', alpha = 0.5) +
  xlab("Parasite") +
  ylab("Growth Rate (Unspecified Units)") +
  theme_bw() +
  coord_flip()
```

  \vspace{0.5cm}   
 
## Observations from BC&P
 
1. We can note that there is substantial 
      variation in the daphnia growth rates
      among the four treatments. 
      
\vspace{0.5cm} 
 
2. We can see that the control treatment
      produces the highest growth rate, about 1.2 mm/day.
      
\vspace{0.5cm}       
 
3. We can see that P. perplexa is closest and
      perhaps lower, M. bicuspidata is next 
      lowest, and P. ramosa is definitely lower.
      
\vspace{0.5cm}       
  
4. We can see that there is likely to be a
      parasite treatment effect overall (question 1),
      and an ordering in the growth rates, with  
      parasites generally driving down the growth rate,
      and P. ramosa < M. bicuspidata < P. perplexa 
      (question 2).
      
\vspace{0.5cm}       
 
5. We can go even further, estimating the average
      growth rate for each treatment 
      (looks like about 1.2 for the control treatment),
      and the difference of the parasite treatments
      from the control treatment (i.e.,the effects of
      each parasite on the growth rate of Daphnia spp.).
      
\vspace{0.5cm} 
 
### 6. Even though BC&P do not do it, we can check some of these comments by calculating a five number summary for the control and each parasite by using the core R function "by()" as follows (first, we will once again view the entire dataset:

  \vspace{0.5cm}  
  
```{r}
# View(daphnia)
?by()

```


```{r}
by(daphnia, daphnia$parasite, summary)
 

``` 
  
   \vspace{0.5cm} 
   
   
  We are going to follow BC&P and use 
  the R function "anova()" in the sections
  on the analysis of variance (ANOVA) and simple
  linear regression.  The core R contains a 
  resident statistical package, "stats".
  This package contains two analysis of variance
  functions "aov()" and "anova()".  The more 
  general and expandable of these is "anova()"
  and BC&P introduce it because they make extensive
  use of it in:
  
\vspace{0.5cm} 
  
    Chapter 6: Advancing Your Statistics in R and
    Chapter 7: Getting Started with Generalized Linear Models. 
    
\vspace{0.5cm} 
 
  As this is an introductory course in 
  statistics, we will not study these methods even
  though they are  useful and frequently used in 
  scientific investigations.  Please feel free
  to read these chapters and create the scripts the
  authors prompt you to write.  Our text and several other
  introductory statistical texts cover two-way ANOVA and
  reading that in conjunction with BC&P will be helpful
  in studying these techniques.  Both of these topics
  are covered in depth in Q SCI 482 and 483.
  
\vspace{0.5cm} 
  
  We will follow BC&P and use
  "anova()" until the end of our analysis
  where we will look at using Tukey's HSD
  (Honestly Significant Difference test) to examine
  which pair(s) of population means are different
  after completing an ANOVA and finding that we
  reject the null hypothesis.  Here, using "aov()"
  along with a simple default "plot()" function
  will easily get us the results we what.
  Both"anova()" and "aov()" offer the ability to 
  perform Tukey's test, but sometimes the less
  expansive function is just easier to use to 
  get the job done. That is a good lesson to remember.
 
  \vspace{0.5cm}   

  
### 7. Compare information on "anova()" and "aov()" by asking for help.
  
\vspace{0.5cm} 
  
```{r}
# ?anova()
?aov()
```

  \vspace{0.5cm} 
  
 
  To perform a one-way ANOVA, we use "anova()"
  along with the function "lm()" which
  constructs a linear model we use as an argument
  in "anova()".  Together with "summary()", these other
  two functions will provide the values we will
  need to perofm the ANVOA on the Daphnia data.
  
\vspace{0.5cm} 
 
  The ANVOA tests the null hypothesis that
  all of the groups come from populations
  that have the same mean.  In this case the 
  four groups are the control and the three 
  parasites that infect the Daphnia. In general,
  we refer to these groups as "treatments".  In
  our experiment, we would call these groups
  the "parasite treatment".
 
   \vspace{0.5cm} 
 
  The test statistic has an F distribution. 
  Its value quantifies the ratio of the 
  between-group variance (between the control and 
  three parasite groups) to the within-group 
  variance (the variance within the control data 
  and within the data for each parasite group).
 
  
\vspace{0.25cm} 
  
 
  We will now perform the ANOVA.  First, we
  estimate a linear model from our data that proposes
  a relationship between the growth rate of Daphnia
  and their infection with parasites. Then,
  check to see if the assumptions of the ANOVA
  are at least approximately met.
 
  
\vspace{0.25cm} 
  
### 8. Fit a linear model for the dependence of growth.rate on parasite.
  
\vspace{0.5cm} 
  
```{r}
model_grow <- lm(growth.rate ~ parasite, data = daphnia)
```

  \vspace{0.5cm}   
  
  Let's search for outliers and then plot
  the box plots.  Notice that a user created
  R function has been used in this script.
  I fwe wrote the script in a contiguous manner, 
  the function could have been created early on
  in the larger script and called (used) here and
  anywhere else we might need it.
  
\vspace{0.5cm} 
 
  We will also search for outliers by visually
  examining the veracity of the assumptions 
  for this experiment using the box plots
  we did and the autoplot() function
  in the package "ggfortify".
  
  
\vspace{0.5cm} 
 
  Be sure to spend some time dissecting what the
  following script (code) segment is doing.  
  Do not forget that dissecting code that
  is already written, and that works, is a good 
  way to learn coding yourself in any software package.
 
  \vspace{0.5cm} 
  
```{r}
is_outlier = function(x) {
  return(x < quantile(x,0.25) - 1.5 * IQR(x) |
           x > quantile(x, 0.73) + 1.5 * IQR(x))
}
daphnia %>%
  group_by(growth.rate) %>%
  mutate(outlier = ifelse(is_outlier(growth.rate), growth.rate, as.numeric(NA))) %>%
  ggplot(., aes(x = factor(parasite), y = growth.rate)) +
  geom_boxplot() +
  geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3)+
  geom_point(size = 1, colour = 'black', alpha = 0.5) +
  xlab("Parasite") +
  ylab("Growth Rate (Unspecified Units)") +
  theme_bw() +
  coord_flip()
```

  
### 9. Plot four diagnostic plots that examine the residuals.
  
```{r}
# install.packages('ggfortify')
library(ggfortify)
autoplot(model_grow, smooth.colour = NA)
```

  \vspace{0.5cm} 
  
 
## Let's Examine the Residual Plots

\vspace{0.35cm} 

Now examine these plots. All of these plots are based on analyzing residuals-errors.

\vspace{0.5cm}   

## Interpretation

\vspace{0.35cm} 
 
The interpretation of these plots is (from BC&P):
 
### 1. Top left. 

\vspace{0.25cm} 

This panel is about the 'systematic part'
of the model; it tells us whether a line is appropriate
to fit to the data. If things have gone wrong, hump-shapes
or valleys will be apparent. These would mean that the
structure of your model was wrong. For example, 
fitting a straight line didn't work.
      
\vspace{0.5cm} 
  
Hence, this plot assesses possible departures
from the linear model we fitted and/or consistently
changing variance. The ideal pattern would be a
      horizontal band of residuals scattered on either
      side of a horizontal line at zero.  Watch out for:
      
\vspace{0.5cm} 
 
a) Curved patterns or other obvious trends in the plot;
  \vspace{0.5cm} 
  
b) Fanning patterns showing the variability consistently increasing or decreasing; and
  \vspace{0.5cm} 
  
c) Outliers that are unusually far above or below the zero line.

  \vspace{0.5cm} 
  
 Many people suggest you look at this to evaluate
      the assumption of equal variance-homo- vs
      heteroskedasticity (unequal variances). But there 
      is a better plot (bottom left) for this.
  \vspace{0.5cm} 
  
### 2. Top right.

  \vspace{0.5cm} 

This Q-Q Plot evaluates the assumption of
normality of the residuals. The dots are the residuals, 
and the dashed line the expectation under the normal
distribution. This is a much better tool than making
a histogram of the residuals, especially with small
sample sizes like less than 100.


\vspace{0.5cm} 
  
### 3. Bottom left.

  \vspace{0.35cm} 


This evaluates the assumption of equal
variance. The y-axis is a standardized (all positive)
indicator of the variation. Linear models assume that
the variance is constant over all predicted
values of the response variable. There should be no
pattern. But there might be one if, for example,
the variance increases with the mean, as it might with
count data (see Chapter 7).


  \vspace{0.5cm} 
  
### 4. Bottom right.

  \vspace{0.35cm} 
  
This evaluates leverage, a tool not
only to detect influential data points, ones that
move the gradient more than might be expected, but 
also to detect outliers. 


   \vspace{0.5cm} 
   
  As BC&P point out, these figures suggest everything is 
  probably fine.  That is, the assumptions of ANOVA
  are met well enough for use to go on with the analysis.
  Never forget that it is up to you as the analyst to
  to justify why the assumptions are met.  Others may wish
  to disagree, so have your agruments we considered.
 
 
\newpage
 
## Using function"anova()"  

  \vspace{0.5cm}  
 
 
  This function, applied to our simple 
  linear model, will provide an answer  
  to our first question posed above:
  
\vspace{0.25cm} 
 
  Is there any effect of our
  treatments (parasite infections and
  no such infection)? In other
  words, do parasites infections affect 
  the growth rate of Daphnia?  
  
\vspace{0.25cm}  

  The simple linear model we fitted allows us to
  propose a relationship between these two variables.
  
\vspace{0.25cm}   
 
  The ANOVA table will help us to analyze the 
  data gathered to determine whether we have
  enough evidence to conclude that a difference
  exists somewhere among the treatments, or if the
  result is due to chance.  
  
\vspace{0.25cm}   
 
  The null hypothesis we are testing is
  that the population mean growth rates for
  each treatment (infection by various parasite 
  species and no parasite) are all equal.  The
  alternative hypothesis is that there is at 
  least one inequality in the mean growth rates.
 
\vspace{0.25cm}   

  
### 10. Calculate the number of observations for each treatment,then print the result.
  
  
 \vspace{0.5cm} 
  
````{r}

observations <- daphnia %>%
  group_by(parasite.name) %>%
    summarise(observations = length(rep))

observations
```
```{r}
observations = 10
print(observations)

```



  \vspace{0.5cm} 
  
### 11. Calculate the number of treatments, then print the result.
  
   \vspace{0.5cm} 
  
```{r}       
library(glue)
treatments = length(unique(daphnia$parasite.name))
glue('# of treatments: {treatments}')
```

  \vspace{0.5cm}   
 
### 12. Calculate the degrees of freedom from the data.
  
   \vspace{0.5cm} 
 
  
                    
   \vspace{0.5cm} 
   
```{r}

between_group_dof = treatments - 1

within_group_dof = observations - treatments

total_dof = observations - 1

glue('Between group: {between_group_dof}, within group: {within_group_dof}, total: {total_dof}')


```
  \vspace{0.5cm} 
  
 
  As we will see when we do linear regression,
  the output from "anova()" will look remarkably
  similar to that for a regression, and it should.
  They are both just linear models. In fact, 
  regression analysis and ANOVA are often 
  described as being two sides of the same
  coin.  They are intimately related.
  
   \vspace{0.5cm} 
  
  From the results of our analysis (the tabular
  output of the anova()), We can see 
  that there is indeed evidence that the 
  parasite treatment, comprising four treatments
  or four levels of manipulation 
  (i. e., a control and three parasites), has
  produced an effect. The F-value (the value
  of the test statistic) far exceeds the critical
  value for our selected alpha value and the
  and the calculated degrees of freedom.
 
 \vspace{0.5cm} 
 
  Notice that the output from "anova()" shows, 
  the between-group variance is large relative
  to the within-group variance.  
  This produces a large F-value (test statistic), 
  and thus a small p-value that allows us to  
  reject the null hypothesis 
  that there are no differences in the 
  population means for the groups.
 
  \vspace{0.5cm} 


  
### 13. Produce a summary
  
\vspace{0.5cm} 
  
```{r}
anova(model_grow)

```

  \vspace{0.5cm} 
  
 
  Let's get more information than just the means.
  This demonstrates that "summarise()"
  can give us a lot of information if 
  we know how to properly request it using "dplyr".
  
   \vspace{0.5cm}  
 
  

  
### 14. Expand what "summarise()" provides, then print results.
  
  \vspace{0.5cm}   
  
```{r}

sumDat<-daphnia %>%
  group_by(parasite) %>%
    summarise(meanGR = mean(growth.rate))
sumDat
```

 \vspace{0.5cm} 
  
 
  Now that we have rejected the null hypothesis
  that there is no difference in the population
  means, it would be natural to ask which of the 
  population means are statistically significantly
  different?  One method of doing this was created
  by the Princeton statistician John W. Tukey. The
  method is called Tukey's Honestly Significant 
  Difference test (Tukey's HSD).  This is one 
  of the things BC&P mean when they write:
 " ...if you've ever heard of the 'multiple
  testing problem', you might be inclined to be
  careful with those p-values. Ask a [statistician not] a 
  knowledgeable friend for a little guidance if that means
  nothing to you."  We will replace the graph that BC&P 
  prompt you to write a script to produce with a plot
  from the results of Tukey's HSD.  There is nothing
  wrong with what BC&P want you to do, but we will
  do something a bit more sophisticated.
  
  \vspace{0.5cm} 
 
  As an aside: Tukey developed exploratory
  data analysis and created many of the methods
  used in it including the 
  5-number summary. Please spend a few minutes and 
  Google him to see what other statistical methods
  he created and, interesting, what commonly used
  words and phrases he coined.
  
  \vspace{0.25cm} 
 
  It should not surprise you that R has Tukey's HSD
  available ( as mentioned above, it is available in
  both "aov()" and "anova()"). To be able to use 
  Tukey's HSD using 'anova()" for developing the 
  ANOVA table, we need the following libraries
 
  
\vspace{0.25cm} 
  
### Add these new R libraries:agricolae, and rstatix. Then, we can run Tukey HSD test analysis.
  
  
  \vspace{0.5cm}   
  
```{r}
# install.packages("agricolae")
# install.packages("rstatix")
library(agricolae)
library(rstatix)


```

  \vspace{0.5cm} 
  
### 15. Get help
  
  \vspace{0.5cm}   
  
  
```{r}
?tukey_hsd()
```

  \vspace{0.5cm}   
 
  Notice what does tukey_hsd() produces and automatically
  displays. It shows the means that are subtracted
  (Group 2 - Group 1), the difference in those means
  (labeled "estimate"), the lower and upper ends of the
  95% CI (the default) and the adjusted p-value for
  Tukey's HSD.  The order, top to bottom, in the table
  is maintained throughout everything that we will see below.
  That makes interpreting the graphical display that we will
  ultimately produce easy. 
 
   
\vspace{0.25cm} 
  
 
  The results of this analysis imply that while the growth
  rate of infected Daphnia are lower than for noninfected
  Daphnia, the growth rates are statistically significantly 
  lower than that of the Control group for
  those Daphnia infected by Metschnikowia bicuspidata
  
  $$ p = 4.36 \times 10^{-10} $$
  
  The 95\% confidence interval (CI) is \([-0.948, -0.515]\) and 
  
  
  
  \vspace{0.25cm} 
  
  In addition, the decrease in Daphnia growth rates 
  is statistically significantly greater for those
  Daphnia infected by Pasteuria ramosa relative to  
  those infected by Pansporella perplexa
  
  $$ p = 6.08 \times 10^{-8} $$
 The 95\% confidence interval (CI) is \([-0.811, -0.377]\) 
  
 
  
\vspace{0.25cm} 
  
 
  For simplicity in creating a visual display of 
  these results, we will use the core R functions
  and the installed libraries to redo this analysis.
  First, we will rerun our ANOVA using "aov()" and
  display the results.   
  Then we will redo the post hoc analysis using 
  Tukey's HSD.
  
  \vspace{0.25cm} 
 
  In both cases, be sure to compare the results 
  obtained using "anova()" to those obtained using
  "aov()". You should expect them to be the same.
 

  \vspace{0.5cm} 
  
  
```{r}
anova(model_grow)
aov(model_grow)
```
  
\vspace{0.5cm} 
  

  
  Now use the function that performs Tukey's HSD,
  then display the results.
  

  
  Get help

\vspace{0.5cm} 
  
```{r}
?tukey_hsd(model_grow)

```

\vspace{0.5cm} 

### Perform the analysis and display the results in tabular form.

\vspace{0.5cm} 

```{r}
posthoc_analysis <- tukey_hsd(aov(model_grow))
posthoc_analysis
```

  \vspace{0.5cm}   
 
  Notice that the results displayed show the
  pairs of means that are subtracted, the order 
  in which they are subtracted, the resulting 
  difference in the means, the lower and upper
  limits of the 95% CI (the default), and the 
  adjusted p-value from Tukey's HSD test.
  
  Please check the difference calculation
  using the table of means that we created above.
 
  \newpage



  
## Visualization
 
  The following "plot()" function
  produces a plot of the results from
  the function TukeyHSD()" using "aov()".
  
  \vspace{0.5cm}   
  
```{r}
plot(posthoc_analysis)
```

  \vspace{0.5cm}   
 
  The graph displays the 95% CI (the default value)
  about the mean difference in growth rates
  for each treatment group pair.  
  
   \vspace{0.5cm}  
  
  Notice that not every pair of means subtracted is
  labeled.  This is how the person writing the code
  chose to handle creating a legible graph. But,
  the subtracted pairs of means in the graph appear
  from top to bottom in the same order in which
  they appeared in the output of the script:
  "posthoc_analysis = TukeyHSD(Test,'parasite')", 
  above.  So, we can identify the pairs not shown
  on the graph.
  
   \vspace{0.5cm}  
 
  Those CI's that are different from zero (i.e., those 
  CI's that do not cross the dashed vertical  
  line at 0.0) are the ones of interest. The vertical
  line at zero indicates no difference in the means.
 
  \vspace{0.5cm}  
  
  The graph clearly shows those pairs that are the 
  most different (in this case showing a decrease in 
  growth rate) are those whose mean difference are 
  negative in the post hoc analysis tables.  Look at the
  line for Pr-Pp at the bottom of the graph.  You should
  be able to see the upper (-0.377) and lower (-0.811) 
  limits of the CI that is given in the two
  versions of the post hoc analysis table;
  as well as the value for the mean difference
  (-0.594).  Comforting, right?
 
  
 \vspace{0.5cm}  
  
 
  You can export plots to incorporate into the  
  assignment you turn in.  You may have to experiment
  with how to do this with the software you are using.
  We will try to help you with this, but you can always 
  save the plot to your desktop and then print it as 
  a separate page and attach it to your solutions.
  
   \vspace{0.5cm}  
 
  If you are using MS WORD, there are at least three ways 
  you can do this: 
  
   \vspace{0.35cm}  
  
1. You can export this plot for pasting into a WORD document
      as a PDF.  In the lower right pane of RStudio click "Export" 
      then select "Save Plot as PDF".  Select the appropriate
      for the plot. Select a good size (4.00 x 6.00 may be a good
      place to start).  Check your plot in "Preview".  When you 
      think that you have a good plot, select a file name and a 
      directory in which to save it.  In WORD, select the Insert tab.
      In the Text group, click Object.  In the Object dialog box,
      click Create from File.  Find the file using Browse, then 
      click Ok.
      
\vspace{0.5cm}  
  
2. You can also save the plot as a metafile. To do this, 
      select "Export", then "copy to Clipboard", 
      then select "Copy as Metafile".
      Make the plot have a height of about 600,
      but you should experiment with this setting to 
      get a good quality graph that clearly displays
      the relationships in the plot.
  
   \vspace{0.5cm}  
   
3. You can also save your plot as a PDF by writing code.
      If we want to save the above plot as a PDF we can add
      the code:
      
\vspace{0.35cm}  
 
        pdf("Post-Hoc Analysis Plot")
        plot(posthoc_analysis)
        dev.off()
        
 \vspace{0.35cm}          
 
  This will save the PDF to the working directory you set 
  up or are working in.
  NOTE: You MUST write "dev.off()" at the end of the code 
  or your computer will keeping trying to save as a PDF 
  until R eventually terminates the session; having
  reached a set time limit.
 
   \vspace{0.5cm}  

  
 
### Now draw the conclusions from the analysis and state them within the context of the problem. Conclusions are written for the indented audience. Write it so that that audience understands what was done and what the conclusions mean in a way meaningful to them.
  
   \vspace{0.5cm}  
  
 
  There are many ways of writing conclusions from
  statistical analysis of data, but one is as follows.
  
   \vspace{0.35cm}  
 
  A one-way analysis of variance was performed to
  to evaluate if parasitic infections affected 
  growth rates in Daphnia spp. for four different
  parasite treatment groups, Control ( no parasitic
  infection), Metschnikowia bicuspidata, 
  Pansporella perplexa, and Pasteuria ramosa. 
  Each group consisted of 10 observations.
  
   \vspace{0.5cm}  
  
  The mean growth rate +/- SEM
  for the Control group was $$ 1.21 \pm 0.0361 $$  
  
This decreased, relative to the Control for Metschnikowia bicuspidata to $$ 0.801 \pm 0.0683 $$
For Pansporella perplexa:  $$1.08 \pm 0.0568 $$ 
For Pasteuria ramosa: $$ 0.482 \pm 0.0613 $$
   
  
   \vspace{0.5cm}  
 
  Daphnia growth rates were statistically sign#ificantly 
  different between treatment groups (parasitic infections),
  assuming an alpha value of 0.05, F = 2.85, versus an
  F(3,36) = 32.33, p = 2.57e-10.
  
   \vspace{0.5cm}  
 
  Post hoc testing analysis using Tukey's Honestly 
  Significant Difference revealed that 
  while the growth rate of infected Daphnia
  are lower than for noninfected Daphnia.
  
  \vspace{0.35cm} 
  
  The growth rates are statistically significantly 
  lower than that of the Control group for
  those Daphnia infected by Metschnikowia bicuspidata
  
  $$ p = 5.76 \times 10^{-5} $$
  
  The 95\% confidence interval (CI) is \([-0.629, -0.196]\) and 
  
  \vspace{0.35cm} 
  
  Pasteuria ramosa: $$ p = 4.36 \times 10^{-10} $$
  
  \vspace{0.35cm} 
  
  The 95\% confidence interval (CI) is \([-0.948, -0.515]\).
  

  
   \vspace{0.5cm}  
  
  In addition, the decrease in Daphnia growth rates 
  is statistically significantly greater for those
  Daphnia infected by Pasteuria ramosa relative to  
  those infected by Pansporella perplexa
  
  \vspace{0.35cm} 
  
  $$ p = 6.08 \times 10^{-8} $$
  
  The 95\% confidence interval (CI) is \([-0.811, -0.377]\) 
  
   \vspace{0.5cm}  
 
### Please make sure that you can calculate the differences shown in the "estimate" and diff" columns of the two post hoc tables.The mean values are given in the display of sumDat1.
 
 
 
                                                   