---
title: "Assignment 8 Analysis Key"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


 
## Logistics:
 
  Please read Chapters 5.4 and 5.5 in BC&P
  while doing the "Simple Linear Regression Tutorial Key.r"
  as you read the chapters.  Then do this
  assignment.
  
  
\vspace{0.55cm}
  
 
## Introduction - The Data.

\vspace{0.55cm}
 
  In this assignment we are going to analyze the relationship
  between where people live, based on latitude, and the 
  mortality rate due to malignant melanoma (a serious cancer 
  with a high death rate). The following is from the CDC in 
  July, 2019:
  
  \vspace{0.55cm}
  
  "Based on data from 2012 to 2016, about 77,698 new cases of
  melanoma occurred in the United States each year, including
  45,854 among men and 31,845 among women. The overall incidence
  rate of melanoma was 21.8 per 100,000. The highest incidence 
  rate was among non-Hispanic white males (34.9 per 100,000), 
  and the lowest rate was among black females (0.9 per 100,000)." 

  \vspace{0.55cm}
 
### Q1. Before we begin, state what you think may come from our analysis.  Do you think that there is some relationship between melanoma mortality rates and latitude? What to you think that relationship might be?
 
  \vspace{0.25cm}
 
I think as latitude approaches the equator, melanoma rates will increase.
  

 \vspace{0.75cm}
 
### Description of the Dataset
 
  The data set consists of mortality data due to malignant  
  melanoma of the skin (a skin cancer) of white males 
  during the period 1950 - 1969 for each state in the 
  United States as well as the District of Columbia.
  No data was available for Alaska, Hawaii, Puerto Rico,
  or the Virgin Islands.  It would have been interesting
  to have those four locations included for our analysis.
  
  \vspace{0.55cm}
 
  The data set has four columns:
 
### 1.  "state": 
The by USPS its abbreviation;

\vspace{0.25cm}

### 2. "mortality": 
the annual number of deaths due to melanoma per
               10,000,000;
               
\vspace{0.25cm}               
               
### 3.  "lat": 
the latitude in degrees (estimated at the center of the state);
In the northern hemisphere, as latitude increases as one moves further north (closer to the North Pole)
         
\vspace{0.25cm}
         
### 4. "population": 
the size of the population in millions as of 1965.
 


\vspace{0.55cm}
 
## Source of the data:

\vspace{0.25cm}
 
  U.S. Department of Health, Education, and Welfare (1974).
 
  We are going to investigate the relationship between melanoma
  mortality and latitude for these states. An example of a similar analysis is:
  
  \vspace{0.35cm}
  
     I. K. Crombie. 1979. Variation of Melanoma Incidence with 
    Latitude in North America and Europe. Br. J. Cnacer 40: 774 - 781.
 
     Reference link:
 
     https://www.cdc.gov/cancer/skin/statistics/index.htm  
 
  
  \vspace{0.35cm}
  
  This time the five steps wil not be explicitly identified. 
  See if you can find the five step process being used in 
  what we do below.
 
  
\vspace{0.55cm}
  
 
### Q2. In the space below, please clear the R memory; load the libraries "ggplot2", "dplyr", "ggfortify", "gridExtra", and "readr", then load the dataset, "melanoma.csv" and assign it to anobject named "Melanoma". 

\vspace{0.55cm}
 
  
```{r}
# Clear the memory
rm(list = ls())


  
# Load libraries
library(ggplot2)
library(dplyr)
library(ggfortify)
library(gridExtra)
library(readr)

# Read or upload dataset
melanoma <- read.csv('melanoma.csv')
```
\vspace{0.55cm}
 

  
 
  Now that we have the data in R, let's take a quick look at it. We will
  want to find out such things as: 
 
  1. What does the dataset look like? 
  2. what type of data is it?
  3. What are the column names in the dataset?
  4. How many rows and columns are in the dataset?
  
  \vspace{0.35cm}
 
  Feel free to look at anything else you would like to
  know about the dataset.
  

 \vspace{0.55cm}
 
```{r}
View(melanoma)
str(melanoma)
colnames(melanoma)
nrow(melanoma)
ncol(melanoma)
```
\vspace{0.55cm}
 
```{r}
# Select alpha = 0.05
Alpha_value  = 0.05
```

  
\newpage

## Part 1: Data Exploration

\vspace{0.55cm}
  
 
  As always, the first step in the analysis will be to explore the data
  looking for the structure of the dataset (what is in it), outliers, and 
  if it appears to meet the assumptions of the hypothesis test we will apply
  in analyzing it.  Plotting the data is critical in the exploration as
  are summary statistics.
 
  \vspace{0.55cm}


  
 
### Q1. In the space below, create an R script that produces at least a summary of the number of locations for which data was collected and the mean, median, and IRQ of the mortality rates. Assign these  to objects named "total_Locations", "meanMortality", "medianMortality", and "irqMoratlity", respectively. You might assign the entire script to an object named "sumData".
 
 \vspace{0.55cm}
 
Feel free to calculate any other sample statistics that you
would like. You may wish to review some prior assignments
for ideas for what to request and for how to write the script. 
 
  \vspace{0.25cm}
 
  Print a title for your table of summary statistics.  You might
  use "Summary Statistics for Melanoma Mortality Data".

\vspace{0.55cm}

```{r}
total_Locations = nrow(melanoma)
meanMortality = mean(melanoma$mortality)
medianMortality = median(melanoma$mortality)
irqMortality = IQR(melanoma$mortality)
# Standard method we have used before:

library(glue)
glue('# locations = {total_Locations}, Mean mort = {meanMortality}, Median mort = {medianMortality}, IQR = {irqMortality}')

```
\vspace{0.55cm}

### Q2. What is the total number of locations, and the mean, median, IRQ of the mortality rates. What do you notice about the mean and the median?
 
  \vspace{0.55cm}

locations = 49, Mean mort = 152.877551020408, Median mort = 147, IQR = 50

The mean is higher than the median by about 6

 \vspace{0.55cm}

 
 \vspace{0.55cm} 
 
 
### Q3. What is the mean latitude? Place your script and the result in the space below.For ease of later use, assign both the mean latitude and the entire script to an object named "meanLatitude".
 

\vspace{0.55cm} 
  

```{r}
meanLatitude <- mean(melanoma$lat)
meanLatitude
```


  \vspace{0.55cm}
 
###. Create a boxplot and check the data for outliers. Incorporate an identification of any outliers in the boxplot.
 
 \vspace{0.55cm}
  
```{r}
is_outlier  = function(x) {
  return(x < quantile(x,0.25) - 1.5 * IQR(x) |
           x > quantile(x, 0.75) + 1.5 * IQR(x))
}
melanoma %>%
  mutate(outlier = ifelse(is_outlier(mortality), mortality, as.numeric(NA))) %>%
  ggplot(., aes(x = "", y = mortality)) +
  geom_boxplot() +
  geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3)+
  geom_point(size = 1, colour = 'black', alpha = 0.5) +
  xlab("") +
  ylab("Melanoma Mortality Rate \n (per 10,000,000 Population") +
  theme_bw()


```
\vspace{0.55cm}


### Q4. Please summarize what you conclude about the data. Think about things like symmetry and outliers.
 
 \vspace{0.25cm} 

ANSWER HERE

\vspace{0.55cm}

 
 
  Plot the mortality data versus latitude in a scatterplot.
  Use informative axis labels and remember to include the 
  appropriate units on each axis.
  
 

```{r}
x = melanoma$lat
y = melanoma$mortality
xrange <- c(min(x), max(x))
yrange <- c(min(y), max(y))
plot(x,y, xlab = 'Lat (degrees)', ylab = 'Melanoma Deaths / 10,000,000 people')
  

```
\vspace{0.55cm}

```{r}
(unique(melanoma$mortality))
# nrow(unique(melanoma$mortality))
```
 
### Q5. From the scatterplot, what observations do you make about the relationship between melanoma mortality rates and latitude?

\vspace{0.25cm}

Melanoma deaths decrease as latitude increases.

\vspace{0.55cm}
 
### Q6. Is the relationship one-to-one (recall from algebra what that means)?
\vspace{0.25cm}

No, the relationship is not one to one. For one thing, there are more values of mortality than latitude, so we will have to have at least one latitude mapping to multiple mortality rates. 

\vspace{0.55cm}
 
### Q7. Is there much scatter?
\vspace{0.25cm}

No, the relationship looks pretty linear.

\newpage


### Linear Models
 
  We want to decide if the scatter seen in the data could be due to some
  relationship between mortality rate and latitude or chance. We may also
  be interested in estimating the melanoma rate for various latitudes.  In 
  addition, we may also like to estimate the variability that is seen 
  in the scatter plot. How might we do this? That is where linear models 
  (linear in the parameters estimated) play a pivotal role.In the case of our data, the overall relationship is roughly
  linear (a straight line), so we will apply simple linear
  regression.
 
  
\vspace{0.55cm}
  
 
  We will hypothesize that the melanoma mortality rate is a 
  function of latitude, using the variables from the
  Melanoma data frame. That is, we are going model this 
  relationship with a simple linear regression model and
  estimate its parameters. 
  
  \vspace{0.75cm}
 
### Q8. Create a script that does the estimation, save the results as an object, named "model_melanoma_mortality" for later use, and print the results.
  
\vspace{0.55cm}
   
   
```{r}
mort = melanoma$mortality
lat = melanoma$lat
model_melanoma_mortality <- lm(mort ~ lat, data = melanoma)
  
# Print the estimates of the regression coefficients,i.e., the estimated intercept and slope.
  
model_melanoma_mortality

```
\vspace{0.55cm}

### Q9. Write the estimated linear model relationship between melanoma mortality rate and and latitude.
 \vspace{0.25cm}  

  The model is: $$ {\rm MMR} = -5.98 *{\rm Latitude} + 389.19 $$

 \vspace{1cm}
  
## Assumptions of a Linear Regression Model 

\vspace{0.55cm}
  
1. The predicted (response) variable come at random from the sampled 
      population and are independent of one another;
      
\vspace{0.35cm}
 
2. The explanatory variables are constants known without error;

\vspace{0.35cm}
  
3. The errors are independent and identically distributed as a 
      normal distributed with a mean of 0 and constant variance.  This is
      the same as saying that the predicted variable has a normal distribution; and
      
\vspace{0.35cm}      
 
4. The variance of the errors is constant (equal to each other; homogeneity).	 

\vspace{1cm}
 
  

  
 
  Check the assumptions of the linear regression
  model.  Do this by visually examining them using
  the "autoplot()" function in the package "ggfortify".
  
  \vspace{0.55cm}
 
### Q1. Run the script and review the plots, then state wht you conclude form this plot and the rest of our exploration of the data that we did above.
 
 \vspace{0.55cm}
  
```{r}

```
\vspace{0.55cm}


  STATE YOUR CONCLUSIONS HERE

\vspace{0.35cm}
 
 
  Let's check another feature of a simple linear regression equation. 
  In the space below, enter the mean of the latitudes into the linear
  regression model with the estimated values of the slope and intercept
  that were determined above.  See if you can do this easily by
  using some of the objects that we created above.  
  
  \vspace{0.55cm}
 
### Q2. What value do you get for the predicted value?  Please show your script and cut and paste the results in the space below.

ANSWER HERE
  
 \vspace{2cm} 
  
### Q3.  What value does this predicted value approximately equal (Hint: You have already calculated this value)?  It is only approximately equal because we have used estimates of the slope  and intercept that do not include many decimal places. 

ANSWER HERE
 
  
\vspace{0.55cm}
 
```{r}
# Recall the mean latitude "meanlatitude" define an object for the mean mortality rate:



# Print the result:



# Calculate the predicted mortality rate for the mean latitude:



# Print the result:



# Subtract this value from the mean mortality rate estimated eralier:

```

  \vspace{0.55cm}

  You get a result that is close to the mean mortality rate calculated
  from the data.  We know that the regression line goes through the point
  (mean of the response variable, mean of the predicted value) =
  (39.5,152.878).  How about that!!
  
 
 \newpage
 
## Part 2: Analysis

\vspace{0.55cm}
 
  We want to decide if the scatter seen in the data could be due to some
  relationship between mortality rate and latitude or chance. In a simple
  linear model the relationship is determined by two parameters, the slope 
  of the equation and the vertical axis intercept.  We know that we can, if
  the assumptions of linear regression analysis are met, conduct
  statistical tests of two hypotheses; on each parameter.
 
  We will only conduct an hypothesis test for the slope.
 
### Q1. State the null and alternative hypotheses for that test.

\vspace{0.55cm}
 


 
 \vspace{0.55cm}
 
## OLS and ANOVA

\vspace{0.55cm}
 
  Recall that linear regression uses ordinary
  least squares to minimize the sum of squares
  (SS) to estimate the coefficients (parameters)
  in the linear model. If we rearrange the SS, 
  apply some algebra, and recognize that the 
  cross-product SS term equals zero, we arrive 
  at a rearranged equation that states:
 
  $$ \text{Total SS} = \text{Regression SS} + \text{Error SS} $$
 
  \vspace{0.55cm}
  
  We use the ANOVA to analyze the sources of
  the Total SS. Hence, regression and ANOVA are
  intimately linked. ANOVA will provide a frame-
  work for testing hypotheses in linear models. 
 
 \vspace{0.55cm}
 
  The ANOVA table is a classic assessment of 
  the null hypothesis, that there is no relationship
  between latitude content and melanoma mortality rate.
  It presents the F-value, degrees of freedom, and 
  the p-value associated with the explanatory 
  variables in the model (in this case, 
  the model has one explanatory variable, 
  i.e., latitude).
 
  \vspace{0.55cm}

  
 
### Q2. Create the script that produces the ANOVA table using the object we created from estimating the coefficients ofthe linear regression model.  

\vspace{0.55cm}

```{r}

```

 \vspace{0.55cm} 
 
### Q3. Calculate and display the summary( table) that provides the regression analysis.
  
 \vspace{0.55cm}
 
```{r}

```

 \vspace{0.55cm}
 
### Q4. How much of the variability in the data is explained my the relationship estimated by the liner model?  Clearly indicate what you looked at to draw this conclusion.
 
 \vspace{0.55cm}

  Looking at the Multiple R-squared:  0.68, we conclude that the relationship 
  between mortality and latitude described by the linear model explains
  approximately 68% of the variability in the data.

\newpage 
 
## A Digression.
 
  Let's stop for a moment and consider something mentioned in the class notes.
  It was stated that the square of the t-distribution is exactly equal to
  the F-distribution; with the appropriate degrees of freedom taken into account.  
  
  \vspace{0.55cm}
  
  In the ANOVA table we see the expected results of an F-test for an
  F-distributed random variable with (1, n-2) = (1,47)
  degrees of freedom.  The values of that test statistic = 99.8.
  \vspace{0.55cm}
  
 
  In the summary model table we see the results of a t-test for a t-distributed
  random variable with $$n-2 = 47$$ degrees of freedom.  The value of that test
  statistic is -9.99.
  
  \vspace{0.55cm}
 
  Notice that $$ -9.99^2 = 99.9 $$. This means that the F-test in the ANOVA is
  EXACTLY equal to the square of the t-test of the null hypothesis that the 
  slope parameter = 0.  Therefore, the t-test and the F-test will provide
  exactly the same result in terms of the p-value and the decision when
  testing the null hypothesis that the population slope parameter = 0.
  The value of the t-test is in one-tailed tests of hypotheses.
 
  \vspace{0.55cm}

  
 
### Now we will complete the test of the hypotheses:

\vspace{0.55cm}
 
  Null hypothesis: the population slope parameter equals 0
  Alternative hypothesis: the population slope parameter does not equal 0.  
 
\vspace{0.55cm}  
### Q1. Calculate the number of observations, n, directly from the data.
\vspace{0.55cm}
  
```{r} 

Num_obs  = length(Melanoma$lat)
Num_obs
```

  
 
## Calculate the critical value for the F-test.
 
### Q2. Run the scripts below, then in the space provided please cut and paste the alpha value, the numerator and denominator degrees of freedom and the valueof the calculated critical value.
 
  
```{r}
Alpha_value

Num_DOF  = 1
Num_DOF

Den_DOF  = Num_obs-2
Den_DOF

Critical_F_Value = qf(1 - Alpha_value, Num_DOF, Den_DOF)
print(Critical_F_Value)
```
 
\vspace{0.55cm}

  
 
### Q3. Report the F test statistic and compare it to critical value. State the decision rule and your conclusion.
 
\vspace{0.55cm}  
 
ANSWER HERE  
 
\vspace{0.55cm}
  
 
## Calculate the critical value for the t-test.

\vspace{0.35cm}
 
### Q4. Run the scripts below, then in the space provided please cut and paste the alpha value, the degrees of freedom and the value of the calculated critical value.

\vspace{0.55cm}
 
  
```{r}
DOF  = Num_obs-2
DOF
print(Criticial_t_Value <- qt(1-(Alpha_value/2), DOF))
```

 \vspace{0.55cm}
 
 
 ANSWER HERE
 
\vspace{0.55cm}
 
### Q5. Report the t test statistic and compare it to critical value. State the decision rule and your conclusion.

\vspace{0.55cm}
 

  ANSWER HERE
 

\vspace{0.55cm}
  
 
  Please take the time to notice that the decisions of each test resulted
  in the same conclusion.Note also that the square of the critical value of the t-test equals
  that of the F-test, and that the square of the test statistic for the
  t-test equals that of the F-test.
 
  \vspace{0.55cm}

\newpage  
 
## Part 3: Predictions and Confidence Intervals
 
  Consider the confidence intervals for both the model and
  predictions made by using it.Show our estimated model together with the original data and 95% confidence intervals. 
 
  
```{r}
ggplot(Melanoma, aes(x = lat,
                     y = mortality)) +
  geom_point() +
  geom_smooth(method = 'lm', se=TRUE) +
  ggtitle ("Linear Model with 95% CI") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab("Melanoma Morality Rate \n (per 10,000,000 Population)") +
  xlab("Latitude (degrees)")
```

\vspace{0.55cm}  
 
Calculate 95% confidence and prediction intervals. Plot the results.
 
## Compute fitted model output and upper (upr) and lower (lwr) prediction intervals. 
 
 \vspace{0.55cm} 
 
  
```{r}
Mortality_Prediction  = predict(model_melanoma_mortality, interval="prediction")


# Combine the upper and lower prediction intervals with the original data.
combined_function  = cbind(Melanoma, Mortality_Prediction)


Mortality_Prediction   # Print the results for checking 
combined_function      # Print the results for checking
```

  \vspace{0.55cm}
 
## Create a scatterplot of the data, the 95% CI for the model, and the prediction intervals.  The red dashed lines are the upper and lower prediction intervals.
 
 \vspace{0.55cm} 
  
```{r}
ggplot(combined_function, aes(x = lat,
                              y = mortality)) +
  geom_point() +
  geom_line(aes(y=lwr), color = "red", linetype = "dashed")+
  geom_line(aes(y=upr), color = "red", linetype = "dashed")+
  geom_smooth(method = 'lm', se=TRUE) +
  ggtitle ("Linear Model with 95% CI and Prediction Intervals") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab("Melanoma Mortality Rate \n (per 10,000,000 Population)") +
  xlab("Latitude (Degrees)")
```  

\vspace{0.55cm}  
 
### Q1. What do you notice about the prediction intervals?  What warning would you give to someone who blindly decides to use a linear regression model to make predictions; especially for ranges far outside of the range of the original data used to fit the model parameters?
 
 \vspace{0.55cm}

  ANSWER HERE

\vspace{0.55cm} 
  
### Q2. Please comment on the following statement:
 
### Living in higher latitudes prevents a person from getting and dying from melanoma.
  
  \vspace{0.55cm}

 

 ANSWER HERE
 
 \newpage

  
 
## Part 4: What information can we obtain from the model and predicted values?

\vspace{0.55cm}
 
  Knowing how to access to this information is useful if you wish
  to do more than test hypotheses with the linear model, e.g., to
  this information in R.
 
  Check for what information is available for
  use from the linear model fitting.
 
 \vspace{0.55cm}
  
```{r}
names(model_melanoma_mortality)

 
# Display the estimated coefficients in the linear model.
 
model_melanoma_mortality$coefficients

 
# Check for what information is available for use in the object "combined_function".
names(combined_function)
 
# Display the mortality rates estimated by the linear regression model.
combined_function$fit 

 
# Display the residuals between the linear regression model and the data.
model_melanoma_mortality$residuals 
```


   \vspace{0.55cm}                                                                              
 
CONGRATULATIONS!!!

\vspace{0.55cm}
 
 
  In the words of Mel Blanc, who was the voice of all the Warner Brothers and
  Looney Tunes cartoon characters who populated the television of my youth, 
 
                   "That's all folks!"
 
  https://en.wikipedia.org/wiki/Mel_Blanc 















