---
title: "Tutorial on Dplyr"
author: "Simon Hans Edasi"
output: html_document
date: "4/11/2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Libraries to Use

\vspace{0.55cm} 

In this tutorial, we will use two main libraries, and focus mainly on the "dplyr" library.  

## 1. ggplot 

\vspace{0.55cm} 

Useful to visualize data, this designs high-quality and customizable plots. A powerful and flexible tool, ggplot creates a diverse array of visualizations such as scatter plots, line charts, bar plots, histograms, etc.  

\vspace{0.25cm} 

It adopts a layed approach, wherein we build a plot one layer at a time. Each layer comprises of the data, its components such as size, shape; and geometric parts such as bars, points, lines. We can customize it by additing legends, scales, and annotations.

\vspace{0.55cm} 

## 2. dyplr

\vspace{0.55cm} 

It's a helpful library to extract insights from a dataset. Composed of functions that operate on data frames and related data structures, we use to summarize, filter, mutate data, arrange, merge, etc. 

\vspace{0.25cm} 

dplyr is a popular data manipulation package in R that provides a fast and efficient way to perform common data manipulation tasks. The purpose of dplyr is to simplify data manipulation and make it easier for users to extract insights from their data.

\vspace{1cm} 


## Let's Begin Coding 

\vspace{0.55cm} 

After the brief overview of the libraries, we will now deploy them and see how they work. Please make sure to run (execute) the R script segments as you follow along reading and rereading Chapter 3 (pp. 57 - 74) in BC&P.
 
\vspace{0.55cm} 

  

  
```{r}
# Clear the memory
rm(list=ls())

# we will use the functions in these libraries
library(ggplot2) # for high-quality graphs
library(dplyr)   # for data cleaning/manipulation
library(readr)  # to import data from CSV files

```

   \vspace{0.55cm}
  
  Import the compensation data set using 
  the path to the data set that you set up.
  Assign the data to a vector named "compensation".
  Then view the data set to ensure it downloaded
  properly.
  
  \vspace{0.55cm}

```{r}
compensation = read_csv("compensation.csv")
# view(compensation)
head(compensation) # shows the first 6 observations of each variable in the dataset
```
 
   \vspace{0.55cm}
   
   
  Using the data that follows we will do what the
  authors of BC&P do.
  We used it in the Second R Tutorial.
  
  \vspace{0.55cm}

  
  

  
```{r}
# Get the names of the columns in compensation
names(compensation)
```

  \vspace{0.55cm} 
  
  
 "dplyr" always takes and gives a data frame as a result.  The "verbs" of "dplyr" are:
  
 \vspace{0.55cm}
 
 
    ?summary()    Returns the mean, median, IQR, Minimum,
                  and maximum of a data set

    ?select()     Selects columns of a data set

    ?slice()      Selects rows of a data set

    ?filter()     Selects subsets of rows conditional
                  upon values in a column.
             
                 "-" Leave out a column or row, e.g, 
                 -Root leave out the Root column 
                 "$" Refers to a variable inside a data
                 frame, e.g., compensation$Root
             
                 Boolean logical operators that Typically
                 can be used:
                "==" equals, e.g., Fruit == 80
                "!=" does not equal, e.g., Fruit != 80
                "<", ">', "<=", ">="
                "|" logical OR
                "&" logical AND
                "!" logical NOT

    ?arrange()  Sorting rows

     ?mutate()  Creates a new variable in the data frame


\newpage                                   
 
##  Section 3.3 Subsetting a data set
 
                                     
 \vspace{0.5cm}
  
  Select the Fruit column in compensation
  
\vspace{0.55cm}
  
```{r}
select(compensation, Fruit)
```

  \vspace{0.55cm}
  
  Select all columns except Root.
  That is accomplished by "-Root". 
  
 \vspace{0.55cm}
 
```{r}
select(compensation, -Root)

# Grab the second row of compensation data set
slice(compensation,2)

# Grab rows 2,3, and 10 of the compensation data set
slice(compensation,c(2,3,10))
  
```

 \vspace{0.55cm} 
 
  See what "with()" does by finding all
  entries in the compensation data set 
  (data.frame) greater than 80 Kg
  
  
   \vspace{0.55cm}

```{r}
with(compensation, Fruit > 80)
```

 \vspace{0.55cm}  
 
  Find the rows where it is true that Fruit is 
  > 80kg return them as a data frame
  
  \vspace{0.55cm}
   
```{r}
filter(compensation, Fruit > 80)
```
 
 \vspace{0.55cm}
  
  Find the rows where it is true that Fruit is > 80 Kg OR < 20, then return the result as a 
  data frame
  
   \vspace{0.55cm}
  
```{r}
filter(compensation, Fruit > 80 | Fruit < 20)
```
 
 \vspace{0.55cm}
  
  Assign the rows where Fruit is > 80 Kg OR < 20 to
  a variable (object) so the results can be used
  
 \vspace{0.55cm}
 
```{r}
lo_hi_fruit = filter(compensation, Fruit > 80 | Fruit < 20)
```
 \vspace{0.55cm}
  
  Have the object's values printed out to check if we have what we want
  
  \vspace{0.55cm}
  
```{r}
lo_hi_fruit
```
 \vspace{0.55cm}
  
  Putting this all together in one code chunk, we get:
  

   \vspace{0.55cm}
  
  
```{r}
# clear memory
rm(list=ls())

  
#  Call the libraries needed (no need to install...)
library(dplyr)
library(ggplot2)

  
#  Enter the desired data set
 

compensation = read_csv("compensation.csv") 

  
#  Check the data: displays the mean, median, std deviation, and number of obs.
summary(compensation)

  
#  Start working with "dplyr" and its capabilities
  

select(compensation, Fruit)   # Gets the Fruit column
select(compensation, -Root)   # Take Root column out from data rows
  
#  Get the 2nd, 3rd & 10th rows of the data set
slice(compensation, c(2,3,10)) 

  
#  Gets rows for each given condition, and assign them to an object.
  

lo_hi_fruit = filter(compensation, Fruit > 80 | Fruit < 20)

  
#  Run this to see what the above line 'saved' for later use.
  

lo_hi_fruit
```

\newpage                            
 
## Section 3.4 Transforming
 
   \vspace{0.5cm}                          


  
  Prior to the transformation, what
  does compensation look like now?
 
  \vspace{0.55cm} 

```{r}
head(compensation)
```

 \vspace{0.55cm}
 
 
  Use mutate to create a new column that is
  "log(Fruit)" add that as the column "logFruit"
  all of which gets put into the object compensation.
  The original data set is NOT changed by this.  We
  are now using a temporary version of the original data set.
  
   \vspace{0.55cm}
  
```{r}
compensation = mutate(compensation, logFruit = log(Fruit))
```
 
 \vspace{0.55cm}
  
  Display the first 7 rows of the new compensation
  data set including the new column "logFruit"
  
   \vspace{0.55cm}
  
```{r}
head(compensation, 7)
```
`
 \vspace{0.55cm}
  
  Use mutate again to create a column that is
  10 times Fruit and call it "TentimesFruit"
   
   
\vspace{0.55cm}


```{r}
compensation = mutate(compensation, TentimesFruit = 10*(Fruit))
```

 \vspace{0.55cm}
  
  Display the first 6 rows of the new compensation
  data set including the new column "TentimesFruit"
  
\vspace{0.55cm}
  
```{r}
head(compensation)
```

 \newpage
                       
 
## Section 3.5 Sorting

\vspace{0.55cm}
 
                       

```{r}  
 # clear memory
rm(list=ls())

  
#  Start with the original data set
compensation = read_csv("compensation.csv")


 # Display the first 6 rows of the original compensation data set 
head(compensation)

  
#  Arrange the Fruit column values of the data set from lowest to highest (the default)
arrange(compensation, Fruit)

  
# Arrange the Fruit column values of the data set from highest to lowest by using "dec()"
arrange(compensation, desc(Fruit))
```

 \vspace{3cm}                            
 
##  Section 3.6 Mini-summary and two top tips
 
                            
 \vspace{0.55cm}
  
 
  Grab only the subset of the data set 
  for Root values from Fruit > 80 Kg.
  Read code from inner parentheses to 
  outer parentheses.  First "filter()" the
  data to get observations where Fruit
  production is > 80 Kg, then Select
  ONLY the Root column and return it.
  
\vspace{0.55cm}
 
  
```{r}
select(filter(compensation, Fruit > 80), Root)
```
  
 
  \vspace{3cm}
  
  
### Piping
 
  Piping symbolized in "dplyr" as "%>%" which
  means: "put the answer of the left-hand
  command INTO the function on the right."
  This is somewhat like a composite function from
  calculus. When using this command, always start with the data frame.
  
   \vspace{0.25cm}
 
  
  
  Below is the script for this.
  Reading from left to right, top to bottom,the
  command below says: [1] working with the
  compensation data, [2] "filter()" it based on the 
  Fruit column, getting all rows where it is
  TRUE that Fruit > 80 Kg, then pass this 
  data frame to [3] "select()" and return ONLY 
  the Root column as the final data frame.
  This is easier and cleaner than putting
  functions inside of functions, i.e., true
  composite functions.
  
\vspace{0.55cm}
 
  
```{r}
compensation %>%
  filter(Fruit > 80) %>%
  select(Root)

  
#  Assign this to an object "Piped_Compensation" for later use.
  

Piped_Compensation=compensation %>% filter(Fruit > 80) %>% select(Root)
Piped_Compensation
```

 \newpage                                 
 
## Section 3.7

 \vspace{0.55cm}


Summarizing data using "dplyr" Calculating summary statistics about groups of data.
 
                                  
 \vspace{0.25cm}
  
 
### Steps in this process:
  
  1. Declare the data frame and what the grouping variable is.
  2. Provide some kind of math function with which to summarize the
     data (e.g. "mean()" or "sd()").
  3. Provide a meaningful name for the values returned.
  4. Make R use all of this information.
 
  

\newpage                                         
  
## Section 3.7.2

\vspace{0.55cm}
 
### Method 1: Nested functions, no piping
 
                                         
 \vspace{0.55cm}
  
  
  Below is the code that does what we want.
 
  [1] First we declare what we want to "summarise()"
  
   \vspace{0.25cm}
 
  The second line of code has some good 
  stuff on the 'inside'.
  
   \vspace{0.25cm}
 
  [2] The "group_by()" function works with
      the data frame and declares Grazing as 
      our grouping variable. If we have more  
      than one grouping variable, we can add
      them with commas in between.
 
  \vspace{0.25cm}
  
  
  [3] The third line is where we ask for the 
      mean to be calculated for the Fruit column. 
      We can do this, and R knows where to look,
      because the "group_by()" function has
      set it all up. 
      
 \vspace{0.25cm}

  The object "meanFruit" is some
  formatting for the output.
 
  "summarise()" returns the results requested.
  There is no need to ask for it to be displayed.
 
   
```{r}
summarise(
  group_by(compensation, Grazing),
  meanFruit = mean(Fruit))
```

 \vspace{0.25cm}
  
 
  If you want to use the means somewhere else in your code, 
  use the "=" symbol and assign the result to a new object. 
  Call it "mean.fruit" as below.
  
   \vspace{0.55cm}
 
  
```{r}
mean.fruit = summarise(
  group_by(compensation, Grazing),
  meanFruit = mean(Fruit))

  
# Return the result.
mean.fruit
```

\newpage
                                             
 
## Section 3.7.3
 
## Method 2: Piping, no nesting of functions
 
                                             
\vspace{0.55cm}
  
 
  This is perhaps more logical in flow than 
  using nested functions.
  
  Start by declaring the data frame.
  
  \vspace{0.25cm}
 
  [1] Start with the data, divide it into groups,
  \vspace{0.25cm}
  
  [2] calculate the mean of the fruit data in each group,
  \vspace{0.25cm}
  
  [3] "summarise()" the results.
  
  \vspace{0.25cm}
 
  NOTE: We work outside to inside when using piping.
 
  A big difference from nesting is that "summarise()"
  is now third in the list, rather than on the 'outside'.  
 
  That's the order.
   
   \vspace{0.55cm}
   
```{r}
compensation %>%
  group_by(Grazing) %>%
  summarise(meanFruit = mean(Fruit))


  
#Assign this to an object for later use
mean.fruit=compensation %>%
  group_by(Grazing) %>%
  summarise(meanFruit = mean(Fruit))


# Display the result; the mean
mean.fruit
```


 \newpage                                          
 
## Section 3.7.4

 \vspace{0.25cm}
 
  summarizing and Extending Summarization
 
  \vspace{0.55cm}                                          
```{r}

compensation %>%
  group_by (Grazing) %>%
  summarise(
    meanFruit = mean(Fruit),
    sdFruit = sd(Fruit))


  
# Assign this to an object for later use.
Fruit.Summary.Statistics = compensation %>%
  group_by (Grazing) %>%
  summarise(
    meanFruit = mean(Fruit),
    sdFruit = sd(Fruit),
    cvFruit = sdFruit/meanFruit)


  # Display the results.
Fruit.Summary.Statistics
```

  \vspace{0.55cm} 
 
  Create a user defined function to calculate
  the coefficient of variation, "cv".
  Use the classic R function "$" to 
  select the desired variable
  from the data set.  Find the "cv"
  of the total Fruit variable (not
  grouped by Grazing).
  We will also display the results.
 
  \vspace{0.55cm}
  
```{r}
cv = function (data) {
  cv.value = sd(data)/mean(data)
  return (cv.value)
}

cv(compensation$Fruit)

Fruit.Summary = compensation %>%
  group_by (Grazing) 
  
cv = function (data) {
  cv.value = sd(data)/mean(data)
  return (cv.value)
}

Fruit.Summary 

cv(Fruit.Summary$Fruit)
```

\newpage




                                 
 
## Appendix 3b: Advanced "dplyr"
 
   \vspace{0.55cm}                               


  
 
  You can apply transformations to groups of your data.
  You already know how to summarize groups of your data, 
  but this is different. Let's say, for example, you want to
  subtract the mean of each group of your data from 
  each value in the corresponding group. 
 
 
  \vspace{0.25cm}
  
  
  To do this, you first declare the groups, just as before,
  using the "group_by()" function. 
  Then you use the "mutate()" function, which you already know 
  as the function for transforming your data.
  The following code does this, subtracting the
  mean fruit production for each grazing
  treatment from the values for the corresponding 
  grazing treatment, using piping. This results in mean-centered data.
 
  We will also assign the result to an object for later use.
 
  \vspace{0.55cm}
  
```{r}
compensation_mean_centered = compensation %>%
  group_by(Grazing) %>%
  mutate(Fruit_minus_mean = Fruit - mean(Fruit))
```

 \vspace{0.55cm}
  
  Check that the mean of each group of the
  "Fruit_minus_mean" variable is equal to zero
  by displaying the results using piping and 
  "summarise()".
  
   \vspace{0.55cm}
  

```{r}
compensation_mean_centered = compensation %>%
  group_by(Grazing) %>%
  mutate(Fruit_minus_mean = Fruit - mean(Fruit))%>%
  summarise(mean.Fruit.Minus.Mean = mean(Fruit_minus_mean))


  
# Display the result required.
  

compensation_mean_centered
```

  \vspace{0.55cm} 
 
  You can ask for more or less of anything to 
  be done to a group of your data. For example, 
  create a linear model for each group. 
  The new function for this is "do()."
  
   \vspace{0.25cm}
   
  Here's how we would "do()" a linear model
  for each grazing treatment:
  We will need to attach the R package
  "broom" to accomplish this. The R function
  "lm()" creates the linear model.
  
   \vspace{0.55cm}
  
```{r}
# library(broom)
# compensation_lms = compensation %>%
#   group_by(Grazing) %>%
#   do(tidy(lm(Fruit ~ Root, data=.)))
```

   \vspace{0.55cm}
 
  Note that we used a function called "tidy()" from 
  the package "broom" to "tidy()" up the output of the
  linear model function "lm()". Otherwise, things are
  rather messy.
 
   \vspace{0.55cm}

  
  
  
```{r}
# Display the results of the linear model
# compensation_lms 
```
\vspace{0.55cm}

  Another way to see the results
  NOTE: the "V" is capitalized!!
  
    
```{r}
# Display the results of the linear model
# View(compensation_lms)
```

  
  
  \vspace{0.55cm}
  
```{r}
?qf()
?qt()
```


\vspace{0.55cm}
                                                  
 
CONGRATULATIONS!!

\vspace{0.25cm}
 
  You have completed the "dplyr" tutorial and
  now have a better understanding of just
  how powerful this function is for manipulating 
  data.  Most importantly, you have learned the 
  basics of using it. Check out the "dplyr" cheat
  sheet for a summary of all of its capabilities.
 
 \vspace{0.25cm}
 
  You are well on your way to to being able to
  use R to analyze your own data or that in 
  someone's lab.  Now there is nothing stopping 
  you from becoming a proficient user of R for
  your own benefit! 
  
 
                                                  