---
title: "QSCI 482 Lab 2"
author: "Simon Hans Edasi"
date: "`r Sys.Date()`"
output: pdf_document
---


\section{Exercise 1: Rounding Data}
Round operated the way I expected, in that it rounded to the nearest integer. I expected this because I have used it before.

```{r}
library(glue)

newdata <- c(1.0, 1.5, 2.0, 2.5, 3.0, 3.5, 4.0)

glue('Ceiling = {ceiling(newdata)}, Floor = {floor(newdata)}, Rounded = {round(newdata)}')
```

Setting digits to a negative value rounds to the nearest power of ten, while positive values determines the number of decimal points to round


```{r}

round(3.456, digits = -2)
round(3456, digits = -1)
round(3456, digits = -3)
round(3456.789, digits = 2)
round(3456.789, digits = 1)

```

\section{Exercise 2}
Probability of two or more people having the same birthday in our class
```{r}
pbirthday(n = 19, classes = 365, coincident = 2)

```

How many people are needed to have a 50% probability?

```{r}
qbirthday(prob = .50, classes = 365, coincident = 2)
```

How many people are needed to have a 99.9% probability?

```{r}
qbirthday(prob = .999, classes = 365, coincident = 2)
```
We need at least 70 people to have a 99.99% probability of two people sharing a birthday.


\section{Exercise 3}
```{r}
?dchisq()
```



```{r}
x = newdata
curve(dchisq(x, df = 3),from = 0, to = 40)
```


```{r}
curve(dchisq(x, df = 30), from = 0, to = 40)
```

\section{Exercise 4}
### Sampling 5 values 10 times from standard normal

```{r}
samps <- c()
for (i in 1:10){
  x = rnorm(n = 5, mean = 0, sd = 1)
  samps[i] <- mean(x)
}
glue('Mean: {mean(samps[1])}, Variance: {var(samps)}')

```


### Sampling 1000 values 5 times from standard normal

```{r}
print('sampling 1000 values from standard normal')
samps <- c()
for (i in 1:5){
  x = rnorm(n = 1000, mean = 0, sd = 1)
  samps[i] <- mean(x)
}
glue('Mean: {mean(samps[1])}, Variance: {var(samps)}')

```
