---
title: "QSCI 482 Story 3"
author: "Simon Hans Edasi"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\section{Question 1}

```{r}
rm(list = ls())
library(glue)
lengths <- read.csv('DaphneHybrids.csv')$Lengths
samp_n <- length(lengths)
samp_var <- var(lengths)
samp_se <- sqrt(samp_var / samp_n)
samp_dof <- samp_n-1
alpha <- 0.05
delta <- 2
n <- 50
delta = 3
for (i in 1:10){
  t <- qt(1-alpha/2, df = n-1)
  n <- ceiling(  (samp_var*t^2) / delta^2  )
}
glue('choose n = {n} to achieve 95% confidence interval of 3mm')
```


\section{Question 2}
### a. If we want to detect an increase in shell length from 310 to 320 $\mu m$ with a significance level of 0.05 and power of 0.90, what sample size is needed under the new protocol? 

```{r}
# H_0: shell_delta >= 10 microns
# H_A: shell_delta < 10 microns
# One tailed alpha
rm(list = ls())
alpha <- 0.05
power <- 0.90
beta <- 1 - power
delta <- 10     # Want to see an increase from 310 to 320
df <- read.csv('WhiskeyCreekAfter.csv')
samp_var <- var(df$ShellLengths)
samp_n <- length(df$ShellLengths)
n <- 50
for (i in {1:10}){
  t_a <- qt(1-alpha, df = n-1, lower.tail = F)
  t_b <- qt(1-beta, df = n-1, lower.tail = F)
  n <- ceiling(  (samp_var / delta^2) * (t_a + t_b)^2  )
}
glue('to detect growth of at least 10 microns we need a sample size of {n}')




```

### b. What sample size is needed for the same requirements as in (a), except with a power of 0.99?

```{r}
# H_0: shell_delta >= 10 microns
# H_A: shell_delta < 10 microns
# One tailed alpha

alpha <- 0.05
power <- 0.99
beta <- 1 - power
delta <- 10     # Want to see an increase from 310 to 320
n <- 50
for (i in {1:10}){
  t_a <- qt(1-alpha, df = n-1, lower.tail = F)
  t_b <- qt(1-beta, df = n-1, lower.tail = F)
  n <- ceiling(  (samp_var / delta^2) * (t_a + t_b)^2  )
}
glue('to detect growth of at least 10 microns with power 0.99 we need a sample
     size of {n}')
```


### c. What sample size is needed for the same requirements as in (a), except with a significance level of 0.10?

```{r}
# H_0: shell_delta >= 10 microns
# H_A: shell_delta < 10 microns
# One tailed alpha

alpha <- 0.10
power <- 0.90
beta <- 1 - power
delta <- 10     # Want to see an increase from 310 to 320
n <- 50
for (i in {1:10}){
  t_a <- qt(1-alpha, df = n-1, lower.tail = F)
  t_b <- qt(1-beta, df = n-1, lower.tail = F)
  n <- ceiling(  (samp_var / delta^2) * (t_a + t_b)^2  )
}
glue('to detect growth of at least 10 microns with alpha = 0.10 and power = 0.90
     we need a sample size of {n}')
```


### d. What sample size is needed for the same requirements as in (a), if we need only detect an increase in shell length from 310 to 330 $\mu m$

```{r}
# H_0: shell_delta >= 10 microns
# H_A: shell_delta < 10 microns
# One tailed alpha

alpha <- 0.05
power <- 0.90
beta <- 1 - power
delta <- 20     # Want to see an increase from 310 to 330
n <- 50

for (i in {1:10}){
  t_a <- qt(1-alpha, df = n-1, lower.tail = F)
  t_b <- qt(1-beta, df = n-1, lower.tail = F)
  n <- ceiling(  (samp_var / delta^2) * (t_a + t_b)^2  )
}
glue('to detect growth of at least 20 microns with alpha = 0.05 and power = 0.90
     we need a sample size of {n}')
```



\section{Question 3}
### Using the data in Question 2, and the same scenario, with a significance level of 0.05, what is the minimum detectable difference in shell length, that can be detected 90% of the time, if

### a. You take a random sample of the shell lengths of 15 baby oysters?
```{r}
alpha <- 0.05
power <- 0.90
beta <- 1-power
samp_n <- 15    # given sample n
t_a <- qt(1-alpha, df = samp_n -1, lower.tail = F)
t_b <- qt(1-beta, df = samp_n - 1, lower.tail = F)
delta <- sqrt(samp_var / samp_n) * (t_a + t_b)
glue('With a sample size of 15 we can detect a minimum difference of 
     {signif(delta,3)} microns')
```


### b. You take a random sample of the shell lengths of 500 baby oysters?

```{r}
alpha <- 0.05
power <- 0.90
beta <- 1-power
samp_n <- 500    # given sample n
delta <- sqrt(samp_var / samp_n) * (t_a + t_b)
glue('With a sample size of 500 we can detect a minimu difference of 
     {signif(delta,3)} microns')
```

\section{Question 4}
### Consider the data and scenario in Question 2, with a significance level of 0.05. It turns out the new protocol actually increases shell growth more than expected, resulting in mean shell length increasing from 310 to 345 $\mu m$. What is the power to detect this difference if:

### a. You take a random sample of the shell lengths of 20 baby oysters?

```{r}
alpha <- 0.05
delta <- 35
samp_n <- 20
se <- sqrt(samp_var / samp_n)
t_a <- qt(1-alpha, df = samp_n - 1)
t_b <- (delta / se) - t_a
power <-1 - pt(t_b,df = samp_n - 1,lower.tail = F)
glue('estimated power is {signif(power,3)}')

```

### b. You take a random sample of the shell lengths of 40 baby oysters?


```{r}
alpha <- 0.05
delta <- 35
samp_n <- 40
se <- sqrt(samp_var / samp_n)
t_a <- qt(1-alpha, df = samp_n - 1)
t_b <- (delta / se) - t_a
power <-1 - pt(t_b,df = samp_n - 1,lower.tail = F)
glue('estimated power is {signif(power,3)}')

```
