---
title: "Assignment 7"
author: "Simon Hans Edasi"
output: pdf_document
date: "2023-06-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


 NOTE: Throughout this and subsequent analysis sections of 
 the assignments, the spaces for your answers will look
 like the space below.


 You may increase the space by using the
 Enter key on your keyboard.  Be sure to prefix any written answers to 
 questions with the  "#" command to avoid R errors.
 Your reply should have each line prefaced by a "#".
 Please remember to press the ENTER key when 
 you wish to go to the next (or a new) line.  Otherwise, R will just let
 you type on the same line forever!! 

 Please try to keep all of your typing confined to what is
 easily visible in RStudio without having to scroll to the right
 so that your work is easily readable.
 Notice how I have typed my comments and code. 
 I have tried to make it readable without scrolling 
 to the right.

 NOTE:

 Please read Chapters 5.4 and 5.6 in BC&P
 while doing the "ANOVA Tutorial Key.r"
 as you read the chapters.  Then do this
 assignment.
 
 Two excellent general references are:
 
 Jerrold H. Zar.2010 (5th edition). Biostatistical Analysis. Prentice Hall.
 (This is often the text for Q SCI 482.)

 Gerald van Belle, L. D. Fisher, P. J. Heagerty, and T Lumley.
 2004 (2nd edition). Biostatisitcs: A Methodology for the Health Sciences.
 Wiley.

 Both of these books provide a broad coverage of statistics including ANOVA
 and linear regression.
 
 \vspace{2cm}
 
 Introduction -- the Experiment:
 
 A completely randomized experiment was conducted to evaluate
 the effects of different insect sprays on insect prevalence 
 in agricultural plots. You may wish to review the class notes
 for Chapter 10 regarding the definition of a completely randomized
 experimental design.
 
 Agricultural plots were randomly selected and assigned at
 random to receive six different insect spray treatments,
 one of which is the control (no treatment). 
 There are 72 plots in the experiment. Each spray was assigned
 to 12 plots ( 6 x 12 = 60) and the control was assigned to 12 plots.
 In this case the treatment actually consists of treating something.
 The plants are sprayed with an insecticide or no insecticide. 
 The control (no insecticide spray used) is also consider to be
 a treatment. You should also notice that the experiment is well
 replicated (there are 12 replicates for each treatment).
 After the treatment, the average number of insects found on
 the plants is recorded.  We have 12 random samples from 6 
 populations

  
 In order to maintain the selected alpha level while comparing 
 the means of more than two groups, an analysis of variance (ANOVA)
 is used rather than multiple pair-wise t-tests (for details, see
 Zar Chapter 10).

 We are going to use a  one-way ANOVA to compare the means of
 the insect counts in the plots.
 The ANOVA is one-way because we are analyzing the effect of only 
 one factor (the insecticide) on the abundance of insects on the
 plants after the treatment.

 The experimental data collected is found in the data set
 "InsectSprays.csv".

 The data set consists of three columns: 
 "number" identifies the plot number;
 "rep" represents the replicate number;
 "count" records the average number of insects; and
 "spray" identifies the treatment ( A through F).



 In the space below, please clear the R memory; load the libraries
 "ggplot2", "dplyr", "ggfortify", and "readr", then load the dataset,
 and assign it to an object named "Insect_Spray".
 
```{r}
#Clear the memory
rm(list = ls())


##
# Load libraries
library(ggplot2)
library(dplyr)
library(ggfortify)
library(readr)

```

\vspace{0.5cm}

Load the InsectSprays.csv data into R:

```{r}
Insect_Spray <- read.csv("InsectSprays.csv")

```

\vspace{1cm}

 Now that we have the data in R, let's take a quick look at it. We will
 want to find out such things as: 

 [1] What does the dataset look like? 
 [2] what type of data is it?
 [3] What are the column names in the dataset?
 [4] How many rows and columns are in the dataset?

 Feel free to look at anything else you would like to
 know about the dataset.
 
 Please enter your code in the space below.
 
```{r}
# View(Insect_Spray)
str(Insect_Spray)
colnames(Insect_Spray)
nrow(Insect_Spray)
ncol(Insect_Spray)
```
 
\vspace{0.5cm}

 Part 1: Data Exploration
 

 As always, the first step in the analysis will be to explore the data
 looking for the structure of the dataset (what is in it), outliers, and 
 if it appears to meet the assumptions of the hypothesis test we will apply
 in analyzing it.  Plotting the data is critical in the exploration as
 are summary statistics.


\vspace{1cm}


 From the ANOVA Tutorial, here is one way to do this. Run the script.
 
```{r}
by(Insect_Spray, Insect_Spray$spray, summary)
```

\vspace{1cm}

Please try to create another way to do this using "dplyr".

 In the space below, create an R script that produces for each treatment
 a summary of the data including means, medians, IRQ's and anything else
 you may wish to display. You may wish to review Assignment 5 for ideas.

```{r}
Insect_Spray %>%
  group_by(spray) %>%
    summarise(
      total_count = sum(count),
      count_IQR = IQR(count),
      count_min = min(count),
      count_mean = mean(count),
      count_max = max(count),
      
    )
```



\vspace{0.5cm}

 Graph the data in whatever way(s) you find most relevant. 
 Make sure to do a boxplot. 

 For this boxplot you do not have to flip the axes, but you might try
 it both ways just for practice with "ggplot" and its parameters.  
 You might label the x-axis "Spray Treatment" and the 
 y-axis "Number of Insects", but the choice of axis labels is 
 ultimately yours.
 
 Then, going back to previous assignments and tutorials, 
 create some code that will incorporate an identification 
 of potential outliers into the boxplot. If your code identifies
 any outliers, do not forget what you have learned to do when you
 encounter them, but we will accept them as correct for the 
 purposes of this assignment and continue on with the analysis of
 the experiment.

 Please put your script in the space below.

\vspace{0.5cm}

```{r}
# Flip the axes
ggplot(Insect_Spray, aes(x = spray, y = count)) +
geom_boxplot() +
theme_bw() +
coord_flip()

# Don't flip the axes
ggplot(Insect_Spray, aes(x = spray, y = count)) +
geom_boxplot() +
theme_bw() 
# coord_flip()

# Incorporate an identification of any outliers in the boxplot.
is_outlier <- function(x) {
  return(x < quantile(x,0.25) - 1.5 * IQR(x) |
           x > quantile(x, 0.75) + 1.5 * IQR(x))
}
Insect_Spray %>%
  group_by(spray) %>%
  mutate(outlier = ifelse(is_outlier(count), count, as.numeric(NA))) %>%
  ggplot(., aes(x = factor(spray), y = count)) +
  geom_boxplot() +
  geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3)+
  geom_point(size = 1, colour = 'black', alpha = 0.5) +
  xlab('Spray') +
  ylab("# of Insects on Plant") +
  theme_bw() 
  # coord_flip()


```

\vspace{1cm}

 Questions:

 [1] Based strictly on your plots, do you think that the sprays have a
     different effect on the insect count?  If so, which ones?  Which
     spray treatment appears to be the most effective in reducing
     the average insect abundance?
     


 [2] Based strictly on your plots which treatment do you think is the
     control (no insecticide applied)?  Why do you conclude this?




 [3] Does your outlier analysis identify any outliers? If so, where?
 

 Please place your answers in the space below.
 
 \vspace{1cm}
 
# Spray C, D, and E seem to have a significant effect on the insects on plants.
# Spray A and B have a reduced variance, but doesn't look significant

# I think spray F is the control. It has the highest number of insects
# and the highest variance of data.

# yes, in sprays C and D there are outliers


\vspace{1cm}

   The ANOVA assumes:

  1. The data consists of independent random samples;
  2. There should be no significant outliers;
  3. The data should be approximately normally distributed; and
  4. The treatment variances are equal.


\vspace{0.5cm}


 Questions:

 [4] Based on the plots, which variable is the dependent variable
     and which is the independent variable? Justify your answers.

 [5] Is assumptions 1 met? Why or why not?
    
 [6] Based on the plots, explicitly state which of the remaining
     assumptions are met and which are not. 
     Does your answer to this cause any concern regarding using
     the ANOVA? If so, why or why not?


 Please place your answers to these questions in the space below.

 \vspace{1cm}

# The dependent variable is the spray A - F, and the independent
# variable is the count of insects found on plants after spraying

# No, the assumption is not met becasue we have found outliers in spray
# C and D

# The data do appear normally distributed from the boxplots

# The variances could potentially be unequal, as the data are spread more
# in boxplots with higher values of counts of insects. 
# These could cause concern for the ANOVA test as the variances are 
# unequal

 \vspace{2cm}
 
    Part 2: ANOVA Test
  

 We are going to use a one-way ANOVA to test whether the sprays have
 the same or different effects as insecticides.

 Assume we select alpha to equal 0.05. 

 We will apply the five step procedure for hypoyhesis testing 
 in our analysis.  As a reminder, those steps, in slightly 
 modified form, are:

 Step 1: State you null and alternative hypothesis. 

 Step 2: Verify that all assumptions necessary for the analysis
         have been met or that we have stated reasons why we can
         use the procedure even if they have not been met.  If met,
         calculate the appropriate test statistic.  We have done 
         some of this and we will do some more.  We will use R
         to calculate the test statistic.

 Step 3: Find the critical value using the appropriate degrees
         of freedom are for the F distribution (in the case of ANOVA).
         Hint: for the F critical value, with k = number of treatments
         and n = number of observations, the numerator degrees of
         freedom is k-1 and k(n-1) is the denominator degrees of 
         freedom. 
         Remember, the F statistic is in the right tail.

 Step 4: Report the F statistic and compare it to critical value.
         State the decision rule.

 Step 5: State your conclusion in the context of the problem.





 Question (Step 1):

 [7] state the null and alternative hypotheses.

 Please place your answer in the space below.


Answer (Step 1):

# H_o: there is no difference in mean insect count between spray treatments
# H_a: There is at least one difference in means between spray treatments

\vspace{2cm}

Step 2: We have done some of this and we will
         do some more below.  We will use R
         to calculate the test statistic.





 Define the linear model and assign it to an
 object named "model_spray", then display the results.

 Once we have doen this, we can create and display
 some diagnostic plots to further evaluate the validity
 of the assumptions for the ANOVA.

```{r}
model_spray <- lm(count ~ spray, data = Insect_Spray)
model_spray

##
# Plot four diagnostic plots that
# examine the residuals.
##

autoplot(model_spray, smooth.colour = NA)
```

\vspace{0.5cm}

 Now examine these plots.
 All of these plots are based on analyzing 
 residuals-errors.  

 Based on what we learned about these four plots 
 in the ANOVA Tutorial, an examination of them 
 shows the some assumptions for the ANOVA 
 are not comfortably met.

 Note: In the wild (the real world) we may opt to
 use a nonparametric procedure (one the does not
 assume anything about the distribution of the underlying
 random variables) to avoid these issues.  An example is
 the Kruskal-Wallis test, which is a nonparametric version
 of the ANOVA. All of these considerations are beyond
 the scope of our class.

 We will rely upon the robustness of the ANOVA, 
 especially since the sample sizes are equal,
 to assume that we may validly use the ANOVA to 
 analyze the experimental data.  Welcome to the
 real world!

\vspace{1cm}

 Step 3: Find the critical value using the appropriate degrees
         of freedom are for the F distribution (in the case of ANOVA).
         Hint: for the F critical value, with k = number of treatments
         and n = number of observations, the numerator degrees of
         freedom is k-1 and k(n-1) is the denominator degrees of 
         freedom. 
         Remember, the F statistic is in the right tail.


\vspace{0.5cm}


 Calculate the number of
 observations (sample size, = n) for each treatment,
 then print the result.

```{r}

Num_obs <- Insect_Spray %>%
  group_by (rep) %>%
  summarise(n())%>%
  nrow()

```

\vspace{0.5cm}

In the space below, please display the result of running
the above script.  Make sure to manually check that these are correct.

```{r}
Num_obs
```

\vspace{0.5cm}

Calculate the number of treatments (= k), then
print the result.

```{r}
Num_treatments <- Insect_Spray %>%
  group_by(spray) %>%
  summarise(n()) %>%
  nrow()

```

\vspace{0.5cm}

In the space below, please display the result of running
the above script.  Make sure to manually check that these are correct.

```{r}
Num_treatments
```

\vspace{0.5cm}

We know that the degrees of freedom (DOF) are:
Numerator DOF = Number of treatments - 1
Denominator DOF = Number of treatments * (Number of observations - 1)

\vspace{0.5cm}

```{r}
Num_DOF <- Num_treatments-1

Den_DOF <- Num_treatments*((Num_obs)-1)

```

\vspace{0.5cm}

In the space below, please display the result of running
this script.  Make sure to manually check that these are correct.

```{r}
print(Num_DOF)
print(Den_DOF)
```

\vspace{0.5cm}

In the space below, write R code to declare the alpha value
to be 0.05; assigning this value to the object "Alpha_value".

```{r}
Alpha_value = 0.05


```

\vspace{0.5cm}

Calculate the critical value of F
with the selected alpha and the 
calculated DOF

```{r}

Critical_F_Value <- qf(1-Alpha_value,
                             Num_DOF,Den_DOF)
```

Display the critical value in the space below

```{r}
Critical_F_Value

```


```{r}
# Create ANOVA table and print the results.

print(ANOVA_Table <- anova(model_spray))
```

Display the ANOVA table

```{r}
ANOVA_Table
```

\vspace{1cm}

Step 4: Report the F statistic and compare it to critical value.
         State the decision rule and it conclusion.

 In the space below, please complete this step.
 
# The F statistic is higher than the criticla value, so we reject the null 
# hypothesis that the means are statistically equal.
 
 
\vspace{1cm}

```{r}
# Find the treatment means.

sumDat <- Insect_Spray %>%
  group_by(spray) %>%
  summarise(meanAbundance = mean(count))

```

Display the results
```{r}
sumDat
```

\vspace{1cm}


 Step 5: State your conclusion in the context of the problem.

 Please complete this step in the space below.

# We can conclude that there is a statistically significant difference of mean
# number of insects when sprayed with different pesticides.




\vspace{0.5cm}



 Question 8:  Are you able to conclude which sprays are most effective
              based on the ANOVA output? Explain.

 Please place your answer to this question in the space below.

# Based on the ANOVA output, no. The ANOVA output only gives us a test statistic,
# we need to do more to determine which spray has a different mean.

\vspace{1cm}


 Assume that you do not know about Tukey's HSD test and that 
 you do not insist that the alpha value remains as selected.
 Suggest a procedure, using your arsenal of statistical knowledge
 and tests, which would allow you to come to the conclusion 
 about which spray is the most effective. You can perform as many
 tests as you like. Briefly explain what you would expect the outcome
 of these tests to be.  

# I would run a t-test on each spray against the assumed control F. I expect
# that these tests would show a different mean for each spray.
 
\vspace{1cm}

 Now use Tukey's HSD test to determine which treatment
 is the most effective as an insecticide.
 We know that Tukey's procedure will test the combination
 of all pairs of treatments.

 Given what you have learned about counting in the probability,
 how many such combinations do you expect to be compared?
 No fair looking ahead in this assignment to see the answer!!
 Make use of the hint above!

 In the space below, please enter you answer.

ANSWER HERE


\vspace{1cm}

```{r}
# Add these new R libraries and the 
# Tukey HSD test analysis can be run
# using the following functions.
#
##

library(agricolae)
library(rstatix)

##
# Get help
##

?tukey_hsd()
```

Perform Tukey HSD analysis and display the results

```{r}
tukey_hsd(model_spray)
```

\vspace{1cm}

 Notice what does tukey_hsd() produces and automatically
 displays. It shows the means that are subtracted
 (Group 2 - Group 1), the difference in those means
 (labeled "estimate"), the lower and upper ends of the
 95% CI (the default) and the adjusted p-value for
 Tukey's HSD.  The order, top to bottom, in the table
 is maintained throughout everything that we will see below.
 That make interpreting the graphical display that we will
 ultimately produce easy.  Note that ns = not 
 statistically significant.


\vspace{0.5cm}


 From an examination of this table, what conclusion do you draw
 about the relative efficacy of the treatment pairs
 B and A, F and A, F and B, D and C, E and C, and E and D?

 Please place your answer in the space below.

# B and A are not significantly different
# F and A are not significantly different
# F and B are not significantly different
# D and C are not significantly different
# E and C are not significantly different
# D and E are not significantly different

# none of these pairs have significantly different results between them.


\vspace{1cm}

 For simplicity in creating a visual display of 
 these results, we will use the core R functions
 and the installed libraries to redo this analysis.
 First, we will rerun our ANOVA using "aov()" and
 display the results.   
 Then we will redo the post hoc analysis using 
 Tukey's HSD.

```{r}
Test <- aov(lm(count ~ spray,data = Insect_Spray))

Test
```

\vspace{1cm}

 Now use the function that performs Tukey's HSD,
 then display the results. Please check some of 
 the results under "diff" to be sure you know
 how they are calculated.  Yuo will need to
 use the table of means that we created above.

 Remember the order of the results in this table
 will be the same as the order of the graphical
 results we will produce below

```{r}
posthoc_analysis <- TukeyHSD(Test,'spray')

```

Display the results

```{r}
posthoc_analysis
```

\vspace{1cm}

      Visualization

 The following "plot()" function
 produces a plot of the results from
 the function TukeyHSD()" using "aov()".

```{r}
plot(posthoc_analysis)
```

\vspace{0.5cm}

The graph displays the 95% CI (the default value)
 about the mean difference in growth rates
 for each treatment group pair; in the same order
 in which the differences in pairs appeared in the 
 tables above.

 Those CI's that are different from zero (i.e., those 
 CI's that do not cross the  dashed vertical
 line at 0.0) are the ones of interest.  

 The vertical line at zero indicates no difference
 in the means. Notice that these CI's are for the
 same pairs as the test concluded that there
 was no statistically significant difference in
 treatment effect.

 The graph clearly shows those pairs that are the 
 most different.  notice the result for the pair 
 F-C, which is the horizontal line that is fourth 
 from the bottom.  Earlier we deduced that F
 was the control treatment (no insecticide spray).

 Please review the boxplots we did while examining this
 visualization of the results of Tukey's HSD. You should
 be able to see that they are both showing the same
 conclusions. That should be comforting!  
 
 Hey, maybe this statistics stuff is not all black magic!!
 
 Looking at this result as displayed in the graph, 
 and from the data displayed in the tables above, 
 what do you conclude about the efficacy of C 
 versus that of all of the other insecticides?
 
 Please place your answer in the space below.

# Given that spray C has the lowest mean count, and the highest difference
# of mean abundance between sprays, we can say that C is the most effective

\vspace{2cm}

 The End of Assignment #7.