---
title: "Assignment 5 KEY"
author: ""
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


 
### Assignment 5 Analysis Key


NOTE: Throughout this and subsequent analysis sections of 
the assignments, the spaces for your answers will look
 like the space below.


 Please read Chapters 5.1 and 5.3 in BC&P
 while doing the "t_tests Tutorial Key.r"
 as you read the chapters.  Then do this
 assignment.
 




 In this assignment we are going to test a hypothesis
 using the two sample t-test.  In Assignment 4
 we calculated CIs and if you read the help 
 information about the "t.test()" and the output
 from using that function, you will realize that
 you have already used the two-tailed, one sample
 t-test for means. You may wish to go back and
 quickly review this.

 You have gradually been building your skills in statistics
 and R, so now is the time for you to start applying
 those skills.

 You will get to create more of the code in this
 assignment than in the previous ones by drawing 
 from code we have written in those assignments.
 We will be here to help in the class tutorial sessions 
 if you need it, but first please try to do it on your own.
 As BC&P mention, there are many sources of R code (including
 from Google searches) and it is not uncommon to use R scripts
 that someone else has already created.

 You can do this!




 
 Introduction -- the Experiment:

 A study was conducted to answer the following question:
 does age affect one's ability to balance while concentrating?
 Nine elderly (6 men and 3 women; aged 75 - 85) and eight young men 
 (aged 20 - 30) were subjects in this experiment.  The researchers
 are interested in answering the question: 
 Does age affect one's ability to balance while concentrating?
 
 Each subject stood barefoot on a "force platform" and was asked
 to maintain a stable upright position and to react as quickly as
 possible to an unpredictable noise by pressing a hand held button. 
 The noise came randomly and the subject concentrated on
 reacting as quickly as possible. The platform automatically measured
 how much each subject swayed in square millimeters.

 
 We are going to assume that the samples are from normally distributed
 populations with equal variances.





## The Data

 The data is found in the dataset "balance.csv". It has three columns. 

 The first column identifies the individual subject by a number, 
 the second contains the combined measurement of the 
 movement of the subjects in square millimeters, and the third
 identifies the age category to which the subject belongs.






 In the space below start your script by loading the dataset.  Assign 
 it to an object named "balance".  We will need the following libraries, so
 load them as well: "ggplot2", "dplyr", and "readr".
 
 Don't forget to clear the memory just increase some unwanted
 data remains in the processor's memory.



```{r}
# Clear the memory.



# Load libraries.






# Load the data set "balance.csv" and assign it to 
# an object named "balance".





```



 Now that we have the data in R, let's take a quick look at it. We will
 want to find out such things as: 


 [1] What does the dataset look like? 
 [2] what type of data is it?
 [3] What are the column names in the dataset?
 [4] How many rows and columns are in the dataset?
enumerate

 Feel free to look at anything else you would like to know
 about the dataset.
 
 Please enter your code in the space below.



```{r}

```



### Part 1. 

 As always , the first step in the analysis will be to explore the data
 looking for the structure of the dataset (what is in it), outliers, and 
 if it appears to meet the assumptions of the hypothesis test we will apply
 in analyzing it.  Plotting the data is critical in the exploration as
 are summary statistics.
 




 Start by calculating summary statistics for each age group. These
 statistics are things like the means, medians, CV's, SME's, min,
 max, IQR, etc.

 In addition to the t-test tutorial, "t_test TutorialKey.r",
 you might try looking back at Assignment 2, where we analyzed 
 data for the length of pike, for some code that 
 will produce this and more. Try doing this in one large script
 using the "dplyr" functions "group_by()" and "summarise()".
 Assign the results to an object in case you want to use it later.
 You may name the object whatever you wish, but you may try "sumData"
 as was done in another assignment.
 
 Take your time and carefully study the previous code. Take care
 to use the correct column names and object names in your code.

 Place your code in the space below.  Make sure to create code that
 will print your summary statistics.


```{r}


```





 Now let's create plots to visually explore the data. We have practiced
 creating several different plots. In exploring the data, we will look 
 for general trends in the data that may give us a clue about [1] whether
 our hypothesis is correct, [2] whether the data has outliers, and [3] if 
 the data satisfies the assumptions of the statistical analysis we propose
 using.  

 We know that the researchers want to analyze their experimental data,
 using the two-sample t-test, to answer the question:
 Does age affect one's ability to balance while concentrating?





## Question 1: What are the assumptions of the two sample t-test? Please list them all in the space below.

ANSWER HERE



## Question 2: What is the population? 

ANSWER HERE


## Question 3: In complete sentences, state the null and alternative hypotheses in terms of the population parameters tested.  Please state them in the space below.  If we assume that muE and muY are the mean movement for the Elderly and Young populations, respectively, state the hypotheses mathematically.

ANSWER HERE


## Question 4: Is this a two-tail or one-tail test.  If the later, which tail are we testing?

ANSWER HERE


## Question 5: What are the appropriate degrees of freedom for the test? Be sure to show how to calculate the degrees of freedom.


ANSWER HERE




 To explore the data, we are going to use three plots to explore the data,
 a histogram, a boxplot, and a Q-Q plot.  
 




 In the space below write an R script to create a histogram for each
 age group.  You might try using "bins=10" and "binwidth=2" in the
 function "geom_histogram()".  Feel free to alter these parameters to
 see if you get anything better than this.  Please provide a label for
 the x-axis of the histograms.  Display one histogram above the other.

```{r}

```



## Question 6: From an examination of the histogram, Which group, if any,  appears to have better balance?  Why do you conclude that? Is this conclusion supported by the summary statistics you calculated?  Be sure to mention the summary statistics in your answer.

ANSWER HERE


 Below is a script that creates a boxplot for each age group and uses 
 the user defined function that we have employed before to check the data 
 for outliers.  Please run it.
 


is_outlier <- function(x) {
  return(x < quantile(x,0.25) - 1.5 * IQR(x) |
           x > quantile(x, 0.73) + 1.5 * IQR(x))
}
balance %>%
  group_by(group) %>%
  mutate(outlier = ifelse(is_outlier(movement), movement, as.numeric(NA))) %>%
  ggplot(., aes(x = factor(group), y = movement)) +
  geom_boxplot() +
  geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3)+
  geom_point(size = 4, colour = 'lightgrey', alpha = 0.5) +
  xlab("Age Group") +
  ylab("Movement (square mm)") +
  theme_bw()

```{r}


```


 Now let's print all of the rows of the modified (mutated) 
 version of "balance" clearly identifying the outliers 
 detected based on using the rule of thumb. Notice the
 addition of the column named "outlier" now present in the 
 data set.  Please compare the values shown with the values 
 printed on the boxplot.  They are the same.

\textbf{Please note you MUST delete the slashes in the code below to get it to run properly.}

balance\$outlier <- ifelse(is\_outlier(balance\$movement), balance\$movement, as.numeric(NA))
print(balance, n=nrow(balance))



```{r}

```


 
## Question 7: Which individual in which group represents a potential outlier and what is its value?  


ANSWER HERE




 We will assume that this value, while different from the others in the 
 group, is valid.





## Question 8; In addition to pointing out a potential outlier, what, if any, affirmations do the the boxplots offer our conclusion about which group has better balance while concentrating ?

ANSWER HERE



 Two of the assumptions of the t-test are that:
 
 [1] the variances are equal and
 [2] that the underlying population is normally distributed.

 Let's check the data to see if these assumptions are met.
 




 Test for equality of variances.
 
 In the space below write an R script that will
 check for the equality of variances in the balance data.

 Review the t-test tutorial for help with this.  Be sure to 
 pay attention to the name of the dataset versus the names of the 
 columns when you use the R function that will do this test.
 
 You might start by getting help on "var.test()".



```{r}


# Get help on "var.test()".
 




# Perform the equality of variance test.




```



## Question 9:  Considering the results of the test for the equality of variances, is there any strong evidence for rejecting the assumption that the variances are equal?

ANSWER HERE







 Check the normality assumption by
 generating a Q-Q Plot.

 Do separate Q-Q Plots for the Elderly and 
 Young groups.

 We will plot these side-by-side
 using the R base "par()" function.
 This will give us a plotting window 
 that consists of one row and two 
 columns.

 We will start by viewing the dataset so that we 
 keep the names of the dataset and columns straight.


```{r}
# Create various Q-Q plots of the data.



```

 Do Q-Q Plot for all of the
 data regardless of the age group.

```{r}

```


## Question 10: Reviewing the Q-Q plots and omitting the one large value in the Elderly group does the data appear to be approximately normally distributed?  Do the boxplots and the summary statistics support this? Are the assumptions of the t-test at least approximately met? Why? 

ANSWER HERE

  


### Part 2

 Apply the two sample t-test. 





 First, read the help file on 
 a two sample t-test.


```{r}  
# ?t.test()
```
  


## Question 11 The researchers performed a power analysis and determined that their alpha level was $0.01$. What does alpha represent? In the space below, please write your answer.

ANSWER HERE

 

 
 Assign the alpha level to an object called "Alpha_value"
 and the degrees of freedom you determined in 
Question 5 to an object named "Degrees_of_freedom".
 
 Please do this in the space below.
 

```{r}

```



 An aside (be sure to run the code (script)) :

 Alternatively, we can calculate the degrees of freedom directly 
 from the data as follows (DOF = Degrees of freedom):


Num_Estimated_Parameters <- 2

Num_Elderly <- length(which(balance$group == 'Elderly'))

Num_Young<- length(which(balance$group == 'Young'))

DOF <- Num_Elderly + Num_Young - Num_Estimated_Parameters

Num_Elderly
Num_Young
DOF

```{r}

```





 
 This is a two tailed test,so we put 
 alpha/2 in each tail.
 
 The resulting positive critical 
 value is found as (getting the
 positive critical value is the
 reason for the 1-(alpha/2) rather
 than using alpha/2).


```{r}  
# print(Critical_Value <- qt(1-(Alpha_value/2),Degrees_of_freedom))
```
  


 Perform a two sample t-test without and with
 Welch's correction. Recall that Welch's correction
 does not assume the equality of variances.  We will
 do the t-test with and without Welch's correction

 We will assign the result of the analyses to objects
 in case we wish to use them later, then we will print
 the results.
 

```{r}

```



 Note in the two outputs for the t-test, the one without
 the Welch correct is called the Two Sample t-test while
 the  one with the Welch corrected is referred to as
 Welch Two sample t-test.  You will not discuss why the 
 degrees of freedom for the Welch test is different from
 the 15 we determined above.

## Question 12: For the test with and without Welch's correction: [a] What is the critical value? [b] What is the value of the test statistic, the associated p-value (include the df output), and the 95% confidence interval?   


ANSWER HERE

## Question 13: [a] What is the decision rule for the test? [b] Within the context of the problem, conclude whether we reject or fail to reject the null hypothesis (state your decision with regard to the hypotheses tested).

ANSWER HERE


  
  



### Part 3.

 Is is always smart to check and double check your work. This was done
 in the experiment we are analyzing.

 One of the grad students on this project found a mistake in the power 
 analysis and realized that the initial alpha estimate was too low for
 the sample size. After redoing the power analysis, she finds that the 
 alpha is 0.1.  This analysis is beyond the scope of our course, but it
 will be extensively covered in Q SCI 482 and 483 as well as other UW
 courses (see Jerrold H. Zar. 2010 (5th edition). Biostatistical Analysis.
 Pearson.).  

 Even without studying power analysis, you should be able to answer 
 the following questions:

## Question 14: [a] In this analysis which error was the graduate student concerned about minimizing in her power analysis?  In a complete sentence, please define that error. [b] Imagine yourself to be this graduate student. What has this discovery done to committing a Type I error? Without doing the analysis, what do you think may now happen to the rejection region for the hypothesis test?  Given this, what would you think may now be the conclusion (reject or fail to reject) regarding the hypotheses test?  Will the critical value and/or the test statistic change?  Should the value that does change increase or decrease using this new alpha value?
  
 

ANSWER HERE




 Let's see if her thoughts are verified.  Rerun the two t-tests
 using the code we have created above.  Don't forget what is 
 now true about the alpha level.  Does the critical value change?
 Does the value of the test statistic change?  Please use the
 space below to enter your code for the t-tests.



```{r}


```



 In the space below please cut and paste the the same information
 you displayed in answering Question 11.  Then, in the context of the 
 problem, state your decision regarding the hypotheses tested.



1.75305

	Two Sample t-test

data:  movement by group
t = 2.7988, df = 15, p-value = 0.01349
alternative hypothesis: true difference in means between group Elderly and group Young is not equal to 0
90 percent confidence interval:
  4.187843 18.228823
sample estimates:
mean in group Elderly   mean in group Young 
             26.33333              15.12500 


	Welch Two Sample t-test

data:  movement by group
t = 2.8784, df = 13.497, p-value = 0.01253
alternative hypothesis: true difference in means between group Elderly and group Young is not equal to 0
90 percent confidence interval:
  4.331855 18.084812
sample estimates:
mean in group Elderly   mean in group Young 
             26.33333              15.12500 



 Some final questions.

## Question 14 (these are asked to give you an opportunity to reflect, so please feel free to give your option(s)): [a] Do you feel more or less confident about the researchers' new conclusions? [b] Given what went on in this experiment, what observation(s) would you make about the importance of selecting the alpha level prior to doing the study (in fact some statisticians would say prior to every collecting the first data point)? Please offer your comments in the space below.

ANSWER HERE



Congratulations, you did it! This is the end of assignment 5.








### An Important Final Note
 
 It is always encumbent upon you to checkif all of the assumptions of the procedures you are using are met by the data. If they are not, then you need to explain why what you are doing is valid!
 


 Interpretation of output from BC&P.


 First, let's look at the arguments we give 
 to the t.test() function: we provide
 a 'formula' and a data argument. The data 
 argument specifies the data frame in which 
 the variables reside. We know that. 
 The formula is a representation of our 
 hypothesis: does movement while balancing during
 concentration depend on age? 
 That is how the "~" (called "tilde") makes us read 
 the formula.

 Let us spend a bit of time reviewing the 
 output as a way to reinforce
 aspects of our workflow.  We will call 
 this the anatomy of the R output:

 [1] The first line of the output tells us 
     the statistical test performed: a Welch
     two-sample t-test. This is what we expected,
     apart from the word 'Welch':
     this is a clear indication that R is doing 
     something slightly unexpected, and we should
     not ignore such an alarm bell. We'll cover this later,
     however.

 [2] Next we see that the data that have been used are 
     declared-this is a good way to confirm that
     you've analyzed the data you wanted to analyze.

 [3] The next line provides the traditional t-test 
     test statistic, degrees of freedom, and p-value.
     We assume you know what these mean. 

 [4] However, let us use the next few lines of output
     to make a few clear points about the two sample
     t-test. The output declares the alternative 
     hypothesis to us: that
     the true difference in means is not equal to 0. 
     This should help you understand more of what 
     you've been doing. 
     If the difference between the two means
     is 0 then they are the same, and we have no 
     grounds for rejecting the null hypothesis.

 [5] Next in the output is a 95% confidence interval. 
     This interval is around the difference between the
     two means. Keep in mind that the difference
     would be 0 if the means were the same. 
     The fact that this interval does not include 0 
     provides an answer to our initial question.
     We can conclude that they are probably different. 
     This falls in line with the value of the test 
     statistic, the critical value for our chosen alpha
     level, and associated p-value.

 [6] Finally, the output provides the means in each
     group.

 Back to the word 'Welch'. A quick look in the help files
 (?t.test) or on Wikipedia reveals that this method 
 is named for the statistician who created it. It allows
 the assumption of equal variances to be relaxed.
 While you may have made the effort to assess
 the assumption of equal variances above, and found
 the variances relatively similar, you now know that there
 are options for when this assumption is not met!













