---
title: "T Tests Tutorial"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


 
## Section 5.3 Two sample t-test (Student's t-test)

 




 Create the script.


```{r}
# Clear the memory.





# Load libraries ggplot2, dpylr, readr


```


 Load the data set "ozone.csv" and assign it to 
 an object named "ozone".

 First, some information about the experiment.
 
 We will analyze ozone levels in 
 gardens distributed around a city. The gardens
 were either to the west of the city 
 center or to the east. The data are ozone 
 concentrations in parts per hundred million
 (pphm). Ozone in excess of 8 pphm can damage
 lettuce plants-they bolt and get filled with 
 latex and are yucky to eat.
 We are interested in whether there is a 
 difference in the average ozone concentration
 between gardens in the east and the west.


```{r}
ozone <- read.csv('ozone.csv')

```


 Check the data set using "View()".  
 NOTE: The "V" is capitalized. This
 is the raw data.  We will go on to
 summarize the data.

 Display the names used in the data set
 using "names()".


```{r}
# View(ozone)




 # Look at a horizontal view of the data
 # using "glimpse()".



```


 Plot the data for exploration
 of the central tendency of the
 data from the two gardens using
 a histogram.

 We thus accomplish the process of visualizing
 our hypothesis and evaluating some
 assumptions of a two-sample t-test, all in one script.


```{r}

```




### A digression into outlier detection and box plots of data.





 Plot the data for further
 exploration for outliers,
 errors, etc. using a box plot.


```{r}


```


 Perform an analysis/checking, then plot the data.

 Create code to detect outliers and label them
 on the box plot using a created function, 
 "is_outlier(x)".

 Start with  the data set "ozone", 
 group_by the Ozone variable, then add a 
 variable "outlier" in a call to "mutate()".
 If a value of Ozone is calculated to be an
 outlier, then that value is passed, otherwise
 an NA is returned so that the value is 
 not plotted.

 The results are plotted.  The text values
 are plotted via geom_text and an aesthetic
 label equal to our new variable, outlier.
 The text for the outlier is offset a small
 amount to the right using hjust = -0.3, so
 that the values are next to rather than on 
 top of the outlier points themselves.


```{r}

```



 
 To explore this outlier detection function, we
 will look at a built-in R data set, "mtcars", 
 that is known to have outliers.




Load needed libraries (dplyr and ggplot2)


```{r}


# Get list of built-in R data sets.



# Get the "mtcars" built-in data set.





# View the data.



```

 Apply the outlier detection function to
 the "mtcars" data set and plot the data 
 as a box plot.


```{r}

```


 Notice that in this data set the function detected what
 the rule of thumb detects to be outliers and the code above
 labels them for us in the boxplot.

 Now let's print all of the rows of the modified (mutated) 
 version of "mtcars" clearly identifying the outliers 
 detected based on using the rule of thumb. Notice the
 addition of the column named "outlier" now present in the 
 data set.  Please compare the values shown with the values 
 printed on the boxplot.  They are the same.
 

 That is pretty useful in helping us decide what to do with
 the values identified with our user defined outlier detection
 function, "is_outlier()"!  
 
 Notice also that we have been able to use that function
 somewhere else in our code without having to
 recreate it again.  That is the value of functions; they
 are always available again in our current code (script).


```{r}

```



 End of digression on outliers in and
 box plot creation for exploring data.







### Resume working through Section 5.3: Two sample t-test (Student's t-test)

 Reenter the previous work (from before the digression on outliers). Create the new script.


```{r}
# Clear the memory.




# Load libraries.


```


 Load the data set "ozone.csv" and assign it to 
 an object named "ozone".

 First, some information about the experiment.
 
 We will analyse ozone levels in 
 gardens distributed around a city. The gardens
 were either to the west of the city 
 centre or to the east. The data are ozone 
 concentrations in parts per hundred million
 (pphm). Ozone in excess of 8 pphm can damage
 lettuce plants-they bolt and get filled with 
 latex and are yucky to eat.
 We are interested in whether there is a 
 difference in the average ozone concentration
 between gardens in the east and the west.


```{r}
ozone <- read.csv('ozone.csv')
```


 Check the data set using "View()".  
 NOTE: The "V" is capitalized. This
 is the raw data.  We will go on to
 summarize the data.

 Display the names used in the data set
 using "names()".


```{r}



# Look at a horizontal view of the data using "glimpse()".



```


 Plot the data for exploration
 of the central tendency of the
 data from the two gardens using
 a histogram.

 We thus accomplish the process of visualizing
 our hypothesis and evaluating some
 assumptions of a two-sample t-test, all in one script.


```{r}

```



 Calculate means, standard deviations,
 standard error of the means, and the coefficient
 of variation of the ozone levels in each 
 location. 


```{r}



# Display the results for the ozone levels.


```

 Perform the two-sample t-test.
 Assume alpha = 0.05.




 Test for equality of variances
 even though the procedure we 
 will use, Welch's method, relaxes
 the assumption of equality of variances, 
 we will still test for the validity of
 that assumption.

 The test for equality of variances
 is an F test.




 Get help on "var.test()".
 

```{r}
# Recall what is in the Ozone dataset

```

 Perform the equality of variance test.

```{r}

```

 Check the normality assumption by
 generating a Q-Q Plot.

```{r}


```


 Do Q-Q Plots for East and West
 garden locations separately.

 We will plot these side-by-side
 using the R base "par()" function.
 This will give us a plotting window 
 that consists of one row and two 
 columns.


```{r}

```

 Do Q-Q Plot for all of the
 data regardless of garden location.

```{r}

```

 Do Q-Q Plot using "ggplot" and
 its functions.

```{r}

```


 We are going to perform a
 two sample t-test where
 the null hypothesis is
 that the means are equal and
 the alternative is that they
 are not equal.
 
 First, read the help file on 
 a two sample t-test.


```{r}
# ?t.test()
```


 Select the alpha value and
 calculate the degrees of
 freedom.
 This is a two tailed test,
 so we put alpha/2 in each tail.


```{r}

```

 Display this in another way.

```{r}

```

 
 This is a two tailed test,
 so we put alpha/2 in each tail.
 
 The resulting positive critical 
 value is found as (getting the
 positive critical value is the
 reason for the 1-(alpha/2) rather
 than using alpha/2).


```{r}

```

 Display the critical value
 in another way.

```{r}

```


 Perform a two sample t-test with
 Welch's correction. Notice
 that the results print automatically
 when the script executes.


```{r}

```


 Perform a two sample t-test with
 Welch's correction. Save the result
 for potential use, then we have to ask for
 the results to be printed.


```{r}

```


 Interpretation of output from BC&P.


 First, let's look at the arguments we give 
 to the t.test() function: we provide
 a 'formula' and a data argument. The data 
 argument specifies the data frame in which 
 the variables reside. We know that. 
 The formula is a representation of our 
 hypothesis: do ozone levels (Ozone) vary as  
 a function of location (Garden.location)?
 That is how the ~ makes us read the formula.


 Let us spend a bit of time reviewing the 
 output as a way to reinforce
 aspects of our workflow.  We will call 
 this the anatomy of the R output:

 [1] The first line of the output tells us 
     the statistical test performed: a Welch
     two-sample t-test. This is what we expected,
     apart from the word 'Welch':
     this is a clear indication that R is doing 
     something slightly unexpected, and we should
     not ignore such an alarm bell. We'll cover this later,
     however.

 [2] Next we see that the data that have been used are 
     declared-this is a good way to confirm that
     you've analyzed the data you wanted to analyze.

 [3] The next line provides the traditional t-test 
     test statistic, degrees of freedom, and p-value.
     We assume you know what these mean. 

 [4] However, let us use the next few lines of output
     to make a few clear points about the two sample
     t-test. The output declares the alternative 
     hypothesis to us: that
     the true difference in means is not equal to 0. 
     This should help you understand more of what 
     you've been doing. 
     If the difference between the two means
     is 0 then they are the same, and we have no 
     grounds for rejecting the null hypothesis.

 [5] Next in the output is a 95% confidence interval. 
     This interval is around the difference between the
     two means. Keep in mind that the difference
     would be 0 if the means were the same. 
     The fact that this interval does not include 0 
     provides an answer to our initial question.
     We can conclude that they are probably different. 
     This falls in line with the value of the test 
     statistic, the critical value for our chosen alpha
     level, and associated p-value.

 [6] Finally, the output provides the means in each
     group.

 Back to the word 'Welch'. A quick look in the help files
 (?t.test) or on Wikipedia reveals that this method 
 is named for the statistician who created it. It allows
 the assumption of equal variances to be relaxed.
 While you may have made the effort to assess
 the assumption of equal variances above, and found
 the variances relatively similar, you now know that there
 are options for when this assumption is not met!

## An IMPORTANT NOTE
 
 IT IS ALWAYS ENCUMBENT UPON YOU TO CHECK IF ALL OF THE
 ASSUMPTIONS OF THE PROCEDURES YOU ARE USING ARE
 MET BY THE DATA.  IF THEY ARE NOT, THEN YOU NEED
 TO EXPLAIN WHY WHAT YOU ARE DOING IS VALID!
 




 End of the t_Tests Tutorial!!



